---
title: "AD Project DE"
author: "Alexander Bergendorf"
date: "2023-08-17"
output: html_document
---

# Library Import
```{r}
library(Seurat)
library(mixOmics)
library(tidyverse)
library(orthogene)
library(dittoSeq)
library(openxlsx)
library(RColorBrewer)
library(pheatmap)
library(scales)
library(VennDiagram)
library(grid)

#home_dir <- "C:/Users/aberg/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
home_dir <- "C:/Users/Alexander/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
```

# Mouse DE
## AD (GSE208683)

```{r}
mouse_gse208683 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
```




```{r}

Idents(mouse_gse208683) <- mouse_gse208683@meta.data$phenotype

de <- FindMarkers(mouse_gse208683, ident.1 = "AD", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/mouse_gse208683_AD_markers_microglia.xlsx"), rowNames = TRUE) 
```

```{r}
de %>%
  filter(p_val_adj < 0.05)
```

864 cells

73 differentially expressed genes


## AD (GSE140510)

```{r}
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))
```


```{r}

Idents(mouse_gse140510) <- mouse_gse140510@meta.data$phenotype

de <- FindMarkers(mouse_gse140510, ident.1 = "AD", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/mouse_gse140510_AD_markers_microglia.xlsx"), rowNames = TRUE) 
```

```{r}
de %>%
  filter(p_val_adj < 0.05)
```

2123 cells

193 differentially expressed genes


## C (GSE208683)

```{r}
mouse_gse208683 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
```




```{r}

Idents(mouse_gse208683) <- mouse_gse208683@meta.data$phenotype

de <- FindMarkers(mouse_gse208683, ident.1 = "C", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/mouse_gse208683_C_markers_microglia.xlsx"), rowNames = TRUE) 
```

```{r}
de %>%
  filter(p_val_adj < 0.05)
```

864 cells


## C (GSE140510)

```{r}
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))
```


```{r}

Idents(mouse_gse140510) <- mouse_gse140510@meta.data$phenotype

de <- FindMarkers(mouse_gse140510, ident.1 = "C", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/mouse_gse140510_C_markers_microglia.xlsx"), rowNames = TRUE) 
```

```{r}
de %>%
  filter(p_val_adj < 0.05)
```

2123 cells

## Plotting DEGs
### Finding overlapping DEGs

```{r}
# ------- Helper: find overlapping genes ranked highest in BOTH studies -------
# Each file is already sorted by FindMarkers (p_val ascending), so row index = rank.
# For overlapping genes, sum the two ranks; smallest combined rank = top priority.
top_overlap <- function(df1, df2, n = 15, p_thresh = 0.05) {
  df1 <- df1 %>% filter(p_val_adj < p_thresh)
  df2 <- df2 %>% filter(p_val_adj < p_thresh)

  df1$rank1 <- seq_len(nrow(df1))
  df2$rank2 <- seq_len(nrow(df2))

  merged <- inner_join(df1[, c("genes", "rank1")],
                       df2[, c("genes", "rank2")],
                       by = "genes")

  merged$combined_rank <- merged$rank1 + merged$rank2
  merged <- merged %>% arrange(combined_rank)

  head(merged$genes, n)
}

# ------- Load per-study mouse DE results -------
# GSE208683 AD markers
sheets <- openxlsx::getSheetNames(paste0(home_dir, "/mouse_gse208683_AD_markers_microglia.xlsx"))
mouse_208683_ad <- lapply(sheets, openxlsx::read.xlsx,
                          xlsxFile = paste0(home_dir, "/mouse_gse208683_AD_markers_microglia.xlsx"))[[1]]
colnames(mouse_208683_ad) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")

# GSE140510 AD markers
sheets <- openxlsx::getSheetNames(paste0(home_dir, "/mouse_gse140510_AD_markers_microglia.xlsx"))
mouse_140510_ad <- lapply(sheets, openxlsx::read.xlsx,
                          xlsxFile = paste0(home_dir, "/mouse_gse140510_AD_markers_microglia.xlsx"))[[1]]
colnames(mouse_140510_ad) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")

# GSE208683 C markers
sheets <- openxlsx::getSheetNames(paste0(home_dir, "/mouse_gse208683_C_markers_microglia.xlsx"))
mouse_208683_c <- lapply(sheets, openxlsx::read.xlsx,
                         xlsxFile = paste0(home_dir, "/mouse_gse208683_C_markers_microglia.xlsx"))[[1]]
colnames(mouse_208683_c) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")

# GSE140510 C markers
sheets <- openxlsx::getSheetNames(paste0(home_dir, "/mouse_gse140510_C_markers_microglia.xlsx"))
mouse_140510_c <- lapply(sheets, openxlsx::read.xlsx,
                         xlsxFile = paste0(home_dir, "/mouse_gse140510_C_markers_microglia.xlsx"))[[1]]
colnames(mouse_140510_c) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")



# Top overlapping AD and C genes across the two mouse studies
mouse_ad_top <- top_overlap(mouse_208683_ad, mouse_140510_ad, n = 15)
mouse_c_top  <- top_overlap(mouse_208683_c,  mouse_140510_c,  n = 15)

combined_genes <- c(mouse_ad_top, rev(mouse_c_top))

```

### Heatmap GSE208683

```{r}
mouse_gse208683 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
mouse_gse208683 <- ScaleData(mouse_gse208683, features = rownames(mouse_gse208683), assay = "RNA")

# Order cells: phenotype first, then orig.ident within phenotype
cell_order <- mouse_gse208683@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  arrange(phenotype, orig.ident) %>%
  pull(cell)

# dittoSeq default colors: orange = "#E69F00", blue = "#56B4E9"
ditto_colors <- c("#E69F00", "#56B4E9")

DoHeatmap(mouse_gse208683,
          features = combined_genes,
          cells = cell_order,
          group.by = "phenotype",
          group.colors = ditto_colors,
          assay = "RNA",
          label = FALSE,
          disp.min = -1,
          disp.max = 2,
          raster = FALSE)

 ggsave(filename = paste0(home_dir, "/figures/manuscript/heatmap_overlap_gse208683.pdf"),
       width = 7, height = 5)
```


### Heatmap GSE140510

```{r}
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))
mouse_gse140510 <- ScaleData(mouse_gse140510, features = rownames(mouse_gse140510), assay = "RNA")

cell_order <- mouse_gse140510@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  arrange(phenotype, orig.ident) %>%
  pull(cell)

# dittoSeq default colors: orange = "#E69F00", blue = "#56B4E9"
ditto_colors <- c("#E69F00", "#56B4E9")

DoHeatmap(mouse_gse140510,
          features = combined_genes,
          cells = cell_order,
          group.by = "phenotype",
          group.colors = ditto_colors,
          assay = "RNA",
          label = FALSE,
          disp.min = -1,
          disp.max = 2,
          raster = FALSE)

 ggsave(filename = paste0(home_dir, "/figures/manuscript/heatmap_overlap_gse140510.pdf"),
        width = 7, height = 5)
```


### Heatmap Combined

```{r}
mouse_gse208863 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))

merged <- merge(mouse_gse208863, mouse_gse140510, add.cell.ids = c("gse208863", "gse140510"), project = "merged")

# In Seurat v5, layers are split by default after merge. Join them so the assay is truly merged:
merged <- JoinLayers(merged)

merged@meta.data$study <- if_else(merged@meta.data$orig.ident %in% c("1_HS", "2_HS", "5_HS", "6_HS"), "GSE208683", "GSE140510")
```

```{r}

cell_order <- merged@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  arrange(phenotype, orig.ident) %>%
  pull(cell)

# dittoSeq default colors: orange = "#E69F00", blue = "#56B4E9"
ditto_colors <- c("#E69F00", "#56B4E9")

DoHeatmap(merged,
          features = combined_genes,
          cells = cell_order,
          group.by = "phenotype",
          group.colors = ditto_colors,
          assay = "RNA",
          label = FALSE,
          disp.min = -1,
          disp.max = 2,
          raster = FALSE)

 ggsave(filename = paste0(home_dir, "/figures/manuscript/heatmap_overlap_combined.png"),
        width = 7, height = 5)
```



# Human DE
## AD (GSE175814)

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse175814_microglia_final.RData"))
```



```{r}

Idents(human_gse175814) <- human_gse175814@meta.data$phenotype

de <- FindMarkers(human_gse175814, ident.1 = "AD", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/human_gse175814_AD_markers_microglia.xlsx"), rowNames = TRUE) 
```


```{r}
de %>%
  filter(p_val_adj < 0.05)
```

2380 cells

2924 differentially expressed genes


## AD (GSE147528)

```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse147528_microglia_final.RData"))
```


```{r}

Idents(human_gse147528) <- human_gse147528@meta.data$phenotype

de <- FindMarkers(human_gse147528, ident.1 = "AD", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/human_gse147528_AD_markers_microglia.xlsx"), rowNames = TRUE) 
```

```{r}
de %>%
  filter(p_val_adj < 0.05)
```

2123 cells

1263 differentially expressed genes


## AD (GSE175814)

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse175814_microglia_final.RData"))
```



```{r}

Idents(human_gse175814) <- human_gse175814@meta.data$phenotype

de <- FindMarkers(human_gse175814, ident.1 = "C", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/human_gse175814_C_markers_microglia.xlsx"), rowNames = TRUE) 
```


```{r}
de %>%
  filter(p_val_adj < 0.05)
```

2380 cells

864 differentially expressed genes


## AD (GSE147528)

```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse147528_microglia_final.RData"))
```


```{r}

Idents(human_gse147528) <- human_gse147528@meta.data$phenotype

de <- FindMarkers(human_gse147528, ident.1 = "C", 
                  only.pos = TRUE, assay = "RNA") 
 

# Save data
write.xlsx(de, file = paste0(home_dir, "/human_gse147528_C_markers_microglia.xlsx"), rowNames = TRUE) 
```

```{r}
de %>%
  filter(p_val_adj < 0.05)
```

2123 cells

1263 differentially expressed genes



## Plotting DEGs
### Finding overlapping DEGs

```{r}
# ------- Helper: find overlapping genes ranked highest in BOTH studies -------
# Each file is already sorted by FindMarkers (p_val ascending), so row index = rank.
# For overlapping genes, sum the two ranks; smallest combined rank = top priority.
top_overlap <- function(df1, df2, n = 15, p_thresh = 0.05) {
  df1 <- df1 %>% filter(p_val_adj < p_thresh)
  df2 <- df2 %>% filter(p_val_adj < p_thresh)

  df1$rank1 <- seq_len(nrow(df1))
  df2$rank2 <- seq_len(nrow(df2))

  merged <- inner_join(df1[, c("genes", "rank1")],
                       df2[, c("genes", "rank2")],
                       by = "genes")

  merged$combined_rank <- merged$rank1 + merged$rank2
  merged <- merged %>% arrange(combined_rank)

  head(merged$genes, n)
}

# ------- Load per-study human DE results -------
sheets <- openxlsx::getSheetNames(paste0(home_dir, "/human_gse175814_AD_markers_microglia.xlsx"))
human_175814_ad <- lapply(sheets, openxlsx::read.xlsx,
                          xlsxFile = paste0(home_dir, "/human_gse175814_AD_markers_microglia.xlsx"))[[1]]
colnames(human_175814_ad) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")

sheets <- openxlsx::getSheetNames(paste0(home_dir, "/human_gse147528_AD_markers_microglia.xlsx"))
human_147528_ad <- lapply(sheets, openxlsx::read.xlsx,
                          xlsxFile = paste0(home_dir, "/human_gse147528_AD_markers_microglia.xlsx"))[[1]]
colnames(human_147528_ad) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")

sheets <- openxlsx::getSheetNames(paste0(home_dir, "/human_gse175814_C_markers_microglia.xlsx"))
human_175814_c <- lapply(sheets, openxlsx::read.xlsx,
                         xlsxFile = paste0(home_dir, "/human_gse175814_C_markers_microglia.xlsx"))[[1]]
colnames(human_175814_c) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")

sheets <- openxlsx::getSheetNames(paste0(home_dir, "/human_gse147528_C_markers_microglia.xlsx"))
human_147528_c <- lapply(sheets, openxlsx::read.xlsx,
                         xlsxFile = paste0(home_dir, "/human_gse147528_C_markers_microglia.xlsx"))[[1]]
colnames(human_147528_c) <- c("genes", "p_val", "avg_log2FC", "pct.1", "pct.2", "p_val_adj")


# Uses the same top_overlap() defined in the mouse block
human_ad_top <- top_overlap(human_175814_ad, human_147528_ad, n = 15)
human_c_top  <- top_overlap(human_175814_c,  human_147528_c,  n = 15)

combined_genes <- c(human_ad_top, rev(human_c_top))
```

### Heatmap GSE175814

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse175814_microglia_final.RData"))
human_gse175814 <- ScaleData(human_gse175814, features = rownames(human_gse175814), assay = "RNA")

cell_order <- human_gse175814@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  arrange(phenotype, orig.ident) %>%
  pull(cell)

# dittoSeq default colors: orange = "#E69F00", blue = "#56B4E9"
ditto_colors <- c("#E69F00", "#56B4E9")

DoHeatmap(human_gse175814,
          features = combined_genes,
          cells = cell_order,
          group.by = "phenotype",
          group.colors = ditto_colors,
          assay = "RNA",
          label = FALSE,
          disp.min = -1,
          disp.max = 2,
          raster = FALSE)

 ggsave(filename = paste0(home_dir, "/figures/manuscript/heatmap_overlap_gse175814.pdf"),
        width = 7, height = 5)
```

### Heatmap GSE147528

```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse147528_microglia_final.RData"))
human_gse147528 <- ScaleData(human_gse147528, features = rownames(human_gse147528), assay = "RNA")

cell_order <- human_gse147528@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  arrange(phenotype, orig.ident) %>%
  pull(cell)

# dittoSeq default colors: orange = "#E69F00", blue = "#56B4E9"
ditto_colors <- c("#E69F00", "#56B4E9")

DoHeatmap(human_gse147528,
          features = combined_genes,
          cells = cell_order,
          group.by = "phenotype",
          group.colors = ditto_colors,
          assay = "RNA",
          label = FALSE,
          disp.min = -1,
          disp.max = 2,
          raster = FALSE)

 ggsave(filename = paste0(home_dir, "/figures/manuscript/heatmap_overlap_gse147528.eps"),
        width = 7, height = 5)
```

### Heatmap Combined

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse175814_microglia_final.RData"))
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse147528_microglia_final.RData"))

merged <- merge(human_gse175814, human_gse147528, add.cell.ids = c("gse175814", "gse147528"), project = "merged")

# In Seurat v5, layers are split by default after merge. Join them so the assay is truly merged:
merged <- JoinLayers(merged)

merged@meta.data$study <- if_else(merged@meta.data$orig.ident %in% c("C1", "C2", "AD1", "AD2"), "GSE175814", "GSE147528")

```





```{r}

cell_order <- merged@meta.data %>%
  tibble::rownames_to_column("cell") %>%
  arrange(phenotype, orig.ident) %>%
  pull(cell)

# dittoSeq default colors: orange = "#E69F00", blue = "#56B4E9"
ditto_colors <- c("#E69F00", "#56B4E9")

DoHeatmap(merged,
          features = combined_genes,
          cells = cell_order,
          group.by = "phenotype",
          group.colors = ditto_colors,
          assay = "RNA",
          label = FALSE,
          disp.min = -1,
          disp.max = 2,
          raster = FALSE)

 ggsave(filename = paste0(home_dir, "/figures/manuscript/heatmap_overlap_combined_human.png"),
        width = 7, height = 5)
```

# Comparing Mouse DE and Human DE

## Comparing Differentially Expressed Genes (Mice)

```{r}
de_mouse_140510 <- read.xlsx(paste0(home_dir, "/mouse_gse140510_AD_markers_microglia.xlsx"), rowNames = TRUE)
de_mouse_140510 <- de_mouse_140510 %>%
  filter(p_val_adj < 0.05)
de_mouse_140510
```

```{r}
de_mouse_208683 <- read.xlsx(paste0(home_dir, "/mouse_gse208683_AD_markers_microglia.xlsx"), rowNames = TRUE)
de_mouse_208683 <- de_mouse_208683 %>%
  filter(p_val_adj < 0.05)
de_mouse_208683
```

```{r}
# Find overlapping row names
overlapping_genes <- intersect(rownames(de_mouse_208683), rownames(de_mouse_140510))

# Get the count
length(overlapping_genes)
```

## Comparing Differentially Expressed Genes (Human)

```{r}
de_human_175814 <- read.xlsx(paste0(home_dir, "/human_gse175814_AD_markers_microglia.xlsx"), rowNames = TRUE)
de_human_175814 <- de_human_175814 %>%
  filter(p_val_adj < 0.05)
de_human_175814
```

```{r}
de_human_147528 <- read.xlsx(paste0(home_dir, "/human_gse147528_AD_markers_microglia.xlsx"), rowNames = TRUE)
de_human_147528 <- de_human_147528 %>%
  filter(p_val_adj < 0.05)
de_human_147528
```

```{r}
# Find overlapping row names
overlapping_genes <- intersect(rownames(de_human_175814), rownames(de_human_147528))

# Get the count
length(overlapping_genes)
```

## Finding Similarities Across Species


```{r}
microglia_140510_mouse_ad <- row.names(de_mouse_140510)


ortholog_mat_140510 <- convert_orthologs(gene_df = microglia_140510_mouse_ad,
                                    input_species = "mouse",
                                    output_species = "human",
                                    non121_strategy = "drop_both_species")
```

```{r}
microglia_208683_mouse_ad <- row.names(de_mouse_208683)


ortholog_mat_208683 <- convert_orthologs(gene_df = microglia_208683_mouse_ad,
                                    input_species = "mouse",
                                    output_species = "human",
                                    non121_strategy = "drop_both_species")
```

```{r}
ortholog_mat_140510
ortholog_mat_208683
de_human_147528
de_human_175814
```

```{r}
# Reorder so GSE175814 and GSE147528 are on the left
gene_sets <- list(
  GSE175814 = rownames(de_human_175814),
  GSE147528 = rownames(de_human_147528),
  GSE140510 = rownames(ortholog_mat_140510),
  GSE208683 = rownames(ortholog_mat_208683)
)

venn.plot <- venn.diagram(
  x = gene_sets,
  category.names = names(gene_sets),
  filename = NULL,
  output = TRUE,
  
  fill = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3"),
  alpha = 0.5,
  lwd = 2,
  
  cex = 1.2,
  cat.cex = 1.3,
  cat.fontface = "bold",
  
  main = "Shared Genes Across Studies",
  main.cex = 1.5
)

grid.newpage()
grid.draw(venn.plot)

pdf(file = paste0(home_dir, "/figures/manuscript/overlapping_genes.pdf"), 
    width = 7, height = 7)
grid.draw(venn.plot)
dev.off()

```

# 



# Plotting Cell-type Abundances

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

cell_counts <- read.csv(file = paste0(home_dir, "/cell_counts_final_all_4_studies.csv"))

# Derive Species from the GSE prefix
cell_counts$Species <- ifelse(
  grepl("GSE140510|GSE208683", cell_counts$Sample),
  "Mouse", "Human"
)

# Pivot to long format
cell_counts_long <- cell_counts %>%
  pivot_longer(
    cols = c(Neuron, Astrocyte, Microglia, Oligodendrocyte, OPC, Endothelial, Fibroblast),
    names_to = "Cell_Type",
    values_to = "Count"
  )

cell_counts_long$sample_species <- paste0(cell_counts_long$Sample, "_", cell_counts_long$Species)

# Set sample order: studies grouped as GSE140510, GSE208683, GSE175814, GSE147528
study_order <- c("GSE140510", "GSE208683", "GSE175814", "GSE147528")
sample_order <- cell_counts_long %>%
  mutate(study = sub("_.*", "", Sample)) %>%
  mutate(study = factor(study, levels = study_order)) %>%
  arrange(study, Sample) %>%
  pull(sample_species) %>%
  unique()

cell_counts_long$sample_species <- factor(cell_counts_long$sample_species, levels = sample_order)

# dittoSeq default colors (Okabe-Ito palette)
ditto_colors <- c(
  "Astrocyte"       = "#E69F00",
  "Endothelial"     = "#56B4E9",
  "Microglia"       = "#009E73",
  "Neuron"          = "#F0E442",
  "Oligodendrocyte" = "#0072B2",
  "OPC"             = "#D55E00",
  "Fibroblast"      = "#CC79A7"
)

p <- ggplot(data = cell_counts_long, aes(x = sample_species, y = Count, fill = Cell_Type)) +
  geom_bar(stat = "identity", color = "black", linewidth = 0.15, position = position_stack()) +
  scale_fill_manual(values = ditto_colors) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
    axis.text.y = element_text(size = 16),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    legend.position = "none"
  ) +
  xlab("Sample") +
  ylab("Total Cell Count")

p
ggsave(filename = paste0(home_dir, "/figures/manuscript/cell_counts_all_studies_stacked.pdf"),
       plot = p, width = 12, height = 7)
p
```


```{r}
colnames(cell_counts)
head(cell_counts)
```


# Session Information

```{r}
sessionInfo()
```


