---
title: "AD Human Processing - GSE157827"
author: "Alexander Bergendorf"
date: "2026-03-18"
output: html_document
---

# Library Import
```{r}
library(Seurat) 
library(tidyverse) 
library(patchwork) 
library(dittoSeq) 
library(stringr)
library(gprofiler2)
library(harmony)
library(patchwork)
library(cowplot)
library(openxlsx)
library(MAST)
library(ggrastr)
library(ragg)

home_dir <- "/mnt/vstor/SOM_PATH_DKB50/members/axb1812/AD_Project_Microglia"

```


# Data Import

```{r}
samples <- c(
  "GSM4775561",  # AD1
  "GSM4775562",  # AD2
  "GSM4775563",  # AD4
  "GSM4775564",  # AD5
  "GSM4775565",  # AD6
  "GSM4775566",  # AD8
  "GSM4775567",  # AD9
  "GSM4775568",  # AD10
  "GSM4775569",  # AD13
  "GSM4775570",  # AD19
  "GSM4775571",  # AD20
  "GSM4775572",  # AD21
  "GSM4775573",  # NC3
  "GSM4775574",  # NC7
  "GSM4775575",  # NC11
  "GSM4775576",  # NC12
  "GSM4775577",  # NC14
  "GSM4775578",  # NC15
  "GSM4775579",  # NC16
  "GSM4775580",  # NC17
  "GSM4775581"   # NC18
)


names <- c(
  "AD1", "AD2", "AD4", "AD5", "AD6", "AD8", "AD9", "AD10", "AD13", "AD19", "AD20", "AD21",
  "NC3", "NC7", "NC11", "NC12", "NC14", "NC15", "NC16", "NC17", "NC18"
)


data_list <- list()

# Load in human data
for(i in samples){ 
  data <- Read10X(paste0(home_dir, "/data/GSE157827/", i)) 
  data_list[[i]] <- data
  print(i)
} 
```

```{r}
names(data_list) <- names
```



# Create Seurat object of human orthologs

```{r}
for(i in names){ 
  data_list[[i]] <- CreateSeuratObject(counts = data_list[[i]], project = i)
} 
```



# Filter for orthologs


```{r}
human_ortholog <- readRDS(file = paste0(home_dir, "/human_orthologs_6datasets.RData"))
```



```{r}
# Filter mouse data based on orthologs
for(i in names){
  data_list[[i]] <- data_list[[i]][human_ortholog]
}
```




# Calculating Quality Control Metrics

```{r}
for(i in names){ 
  data_list[[i]][["rna.size"]] <- log10(Matrix::colSums(GetAssayData(data_list[[i]], assay = "RNA", layer = "counts")))
  data_list[[i]][["percent.mt"]] <- PercentageFeatureSet(data_list[[i]], pattern = "^MT-")
}
```

# QC and Visualization


```{r}
data_list_test <- merge(x = data_list[[1]], y = data_list[2:length(data_list)], merge.data = TRUE)
df <- data_list_test@meta.data

# RNA size
p1 <- ggplot(df, aes(x = orig.ident, y = rna.size, fill = orig.ident)) +
  geom_violin() +
  rasterise(geom_jitter(width = 0.25, size = 0.1, alpha = 0.4),
            dpi = 300, dev = "ragg") +
  theme_classic(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Sample", y = "RNA Size")
ggsave(paste0(home_dir, "/Manuscript Figures/rnasize_qc_gse157827.eps"),
       plot = p1, device = "eps",
       width = 6, height = 5, units = "in")

# Percent MT
p2 <- ggplot(df, aes(x = orig.ident, y = percent.mt, fill = orig.ident)) +
  geom_violin() +
  rasterise(geom_jitter(width = 0.25, size = 0.1, alpha = 0.4),
            dpi = 300, dev = "ragg") +
  theme_classic(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Sample", y = "Percent MT")
ggsave(paste0(home_dir, "/Manuscript Figures/mt_qc_gse157827.eps"),
       plot = p2, device = "eps",
       width = 6, height = 5, units = "in")
rm(data_list_test)

```



```{r}
rna_size_min <- list() 
rna_size_max <- list() 

# Calculates 3 MAD from the median rna-size, filters out if outside that range
for(i in names){ 
  rna_size_min[[i]] <- median(data_list[[i]]@meta.data$rna.size) - (3*mad(data_list[[i]]@meta.data$rna.size)) 
  
  rna_size_max[[i]] <- median(data_list[[i]]@meta.data$rna.size) + (3*mad(data_list[[i]]@meta.data$rna.size)) 
} 
```



```{r}
# Save QC metric DF
qc_metrics <- c() 
for(i in names){ 
  preqc_cells <- ncol(data_list[[i]]) 
  qc_metrics <- rbind(qc_metrics, preqc_cells) 
}

colnames(qc_metrics) <- "preqc_cells" 
row.names(qc_metrics) <- NULL 
preqc_metrics <- as.data.frame(qc_metrics)
```



```{r}
# Filter based on QC metrics
for(i in names){ 
  temp_max <- rna_size_max[[i]] 
  temp_min <-  rna_size_min[[i]] 
  data_list[[i]] <- subset(data_list[[i]], subset = percent.mt < 10 & rna.size < temp_max & rna.size > temp_min) 
  
}
```


```{r}
# Finish QC Table
qc_metrics <- c() 
for(i in names){ 
  preqc_cells <- ncol(data_list[[i]]) 
  qc_metrics <- rbind(qc_metrics, preqc_cells) 
}
```



```{r}
colnames(qc_metrics) <- "postqc_cells" 
row.names(qc_metrics) <- NULL 
postqc_metrics <- as.data.frame(qc_metrics)
cellcounts <- cbind(preqc_metrics, postqc_metrics)
write.csv(cellcounts, file = paste0(home_dir, "/results/gse157827_human_cellcounts_ortholog.csv"))
```


```{r}
cellcounts
```




# Check Cell Cycle after clustering

```{r}
# Cell Cycle Scoring Addition

s.genes <- cc.genes$s.genes 
g2m.genes <- cc.genes$g2m.genes 

for(i in names){ 
  data_list[[i]] <- NormalizeData(data_list[[i]]) 
  data_list[[i]] <- CellCycleScoring(data_list[[i]], s.features = s.genes, g2m.features = g2m.genes) 
} 
```

# SCTransform
```{r include=FALSE}
#Applies SCTransform individually to each sample 
data_list <- lapply(X = data_list, FUN = SCTransform,  method ="glmGamPoi", return.only.var.genes = FALSE, vars.to.regress = c("S.Score", "G2M.Score")) 
```


```{r}
# Save data after SCTransform, computationally intensive
#saveRDS(data_list, file = paste0(home_dir, "/sc_human_objects/data_list_ortholog.RData"))
#processed_list <- readRDS(file = paste0(home_dir, "/sc_human_objects/data_list_ortholog.RData"))
```



```{r}
processed_list <- data_list
rm(data_list)
```


# Batch

```{r}
# Select integration features across samples
variable.features <- SelectIntegrationFeatures(object.list = processed_list, nfeatures = 3000)

# Start with the first Seurat object
combined <- processed_list[[1]]

# Create a list with the remaining objects
remaining_objects <- processed_list[-1]

# Optionally remove the original list to free memory
rm(processed_list)
gc()

# Merge objects sequentially while removing them from the list and printing progress
while (length(remaining_objects) > 0) {
  # Merge the first remaining object into 'combined'
  combined <- merge(combined, remaining_objects[[1]])
  
  combined <- JoinLayers(combined, assay = "RNA")

  # Remove the merged object from the list
  remaining_objects[[1]] <- NULL
  
  # Print the number of objects left to merge
  cat("Remaining objects to merge:", length(remaining_objects), "\n")
  
  # Prompt garbage collection to free memory
  gc()
}


# Assigns variable features calculated across samples
VariableFeatures(combined) <- variable.features 
```



```{r}
# Performs PCA
combined <- RunPCA(combined, verbose = FALSE) 
```

```{r}
variance <- (combined@reductions$pca@stdev)^2
variance <- variance / sum(variance)
variance
```


```{r}
# PC Plot before integration occurs
options(repr.plot.height = 12, repr.plot.width = 20) 
p1 <- DimPlot(object = combined, reduction = "pca", pt.size = 2, group.by = "orig.ident", raster = TRUE) 
#p2 <- VlnPlot(object = combined, features = "PC_1", group.by = "orig.ident", pt.size = 0) 
#plot_grid(p1,p2)
ggsave(filename= paste0(home_dir, "/Manuscript Figures/gse157827_pre_batch.pdf"),
       plot = p1, device = cairo_pdf, width = 6, height = 5, units = "in") # name the file as an EPS
```


We should batch correct.

```{r include=FALSE}
# Running Harmony  
options(repr.plot.height = 8, repr.plot.width = 16) 

pdf(NULL)
combined <- RunHarmony(combined, assay.use = "SCT", group.by.vars = "orig.ident", plot_convergence = TRUE) 
dev.off()
```

```{r}
variance <- (data_list@reductions$harmony@stdev)^2
variance <- variance / sum(variance)
variance
```




```{r}
# PC plot after integration occurs
p1 <- DimPlot(object = combined, reduction = "harmony", pt.size = 2, group.by = "orig.ident", raster = TRUE) 

#p2 <- VlnPlot(object = data_list, features = "harmony_1", group.by = "orig.ident", pt.size = 0) 
ggsave(filename= paste0(home_dir, "/Manuscript Figures/gse157827_post_batch.pdf"),
       plot = p1, device = cairo_pdf, width = 6, height = 5, units = "in") # name the file as an EPS
```



```{r}
# Elbowplot to select PCs for clustering
p1 <- ElbowPlot(combined, reduction = "harmony", n = 50) 

ggsave(filename= paste0(home_dir, "/figures/gse157827/elbowplot.png"), width = 9, height = 7) # name the file as an EPS
```

Start with 30 PCS, check separation


# Reducing Dimensionality and Clustering

```{r}
# UMAP
combined <- RunUMAP(combined, reduction = "harmony", dims = 1:30, verbose = FALSE) 

# Find Neighboring cells 
combined <- FindNeighbors(combined, reduction = "harmony", dims = 1:30, verbose = FALSE) 

# knn clustering
combined <- FindClusters(combined, resolution = 1, verbose = FALSE) 
```


```{r}
# Visualize data
p1 <- dittoDimPlot(combined, "seurat_clusters", do.label = TRUE, legend.show = FALSE, labels.size = 3) 
ggsave(filename= paste0(home_dir, "/figures/gse157827/seurat_clusters.png"), width = 9, height = 7) # name the file as an EPS

p1 <- dittoDimPlot(combined, "orig.ident") 
ggsave(filename= paste0(home_dir, "/figures/gse157827/orig.ident.png"), width = 9, height = 7) # name the file as an EPS

p1 <- dittoDimPlot(combined, "percent.mt") 
ggsave(filename= paste0(home_dir, "/figures/gse157827/percent.mt.png"), width = 9, height = 7) # name the file as an EPS

p1 <- dittoDimPlot(combined, "Phase") 
ggsave(filename= paste0(home_dir, "/figures/gse157827/Phase.png"), width = 9, height = 7) # name the file as an EPS

```

```{r}
# Save dataset after clustering
#saveRDS(combined, file = paste0(home_dir, "/sc_human_objects/gse157827_final.RData"))
combined <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse157827_final.RData"))
```

# Visualizing Data


```{r}
# Visualization of some markers
marker_list <- c("GAD1", "NRGN",
             "ANLN", 
             "VCAN", 
             "CABLES1", "AQP4", 
             "VWF", "FLT1",
             "CSF1R", "DCN")
Idents(combined) <- combined@meta.data$seurat_clusters
for(i in marker_list){
  p1 <- dittoDimPlot(combined, var = i, assay = "RNA")
  ggsave(filename = paste0(home_dir, "/figures/gse157827/markers/umap_", i, ".png"),
         plot = p1, width = 9, height = 7)
}

# Dot plot of all markers across clusters
p2 <- DotPlot(combined, features = marker_list, assay = "RNA") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_discrete(limits = rev)
ggsave(filename = paste0(home_dir, "/figures/gse157827/markers/dotplot_markers.png"),
       plot = p2, width = 9, height = 7)
```




# Finding Cluster Identifiers

```{r}
# Get unique identifiers
unique_idents <- as.character(unique(combined@meta.data$seurat_clusters))
unique_idents

# Initialize an empty list to store the results
markers_list <- list()

# Loop over each identifier and run FindMarkers
for (id in unique_idents) {
  markers_list[[id]] <- FindMarkers(
    object = combined,
    ident.1 = id,
    only.pos = TRUE,
    assay = "RNA",
    max.cells.per.ident = 300
  )
}

saveRDS(markers_list, paste0(home_dir, "/results/all_cell_markers_human_ortholog_gse157827.RData"))
```


```{r}
# Save dataset after clustering
#saveRDS(combined, file = paste0(home_dir, "/sc_human_objects/gse157828_final.RData"))
combined <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse157827_final.RData"))
```

```{r}
markers_list <- readRDS(paste0(home_dir, "/results/all_cell_markers_human_ortholog_gse157827.RData"))
# Get counts per cluster from the Seurat object
cluster_counts <- table(combined$seurat_clusters)

# Rename each list element: "clusternumber_clustercount"
for (cl in names(markers_list)) {
  names(markers_list)[names(markers_list) == cl] <- paste0(cl, "_n", cluster_counts[[cl]])
}
```



```{r}
write.xlsx(markers_list, file = paste0(home_dir, "/results/all_cell_markers_human_ortholog_gse157827.xlsx"), rowNames = TRUE)
```

0- Oligodendrocyte
1- Neuron
2- Oligodendrocyte
3- Neuron
4- Astrocyte
5- Oligodendrocyte
6- OPC
7- Neuron
8- Neuron
9- Microglia
10- Neuron
11- Neuron
12- Astrocyte
13- Neuron
14- Neuron
15- Neuron
16- Neuron
17- Neuron
18- OPC
19- Neuron
20- Endothelial 
21- Doublet
22- Neuron
23- Neuron
24- Neuron
25- Doublet
26- Doublet
27- Doublet
28- Neuron
29- Doublet
30- OPC
31- Doublet
32- Doublet


# Assign Idents

```{r}
combined <- subset(combined, subset = seurat_clusters %in% c(21, 25, 26, 27, 29, 31, 32), invert = TRUE)
# Summary Idents
Idents(combined) <- combined@meta.data$seurat_clusters 
new.cluster.ids <- c("Oligodendrocyte", "Neuron", "Oligodendrocyte", "Neuron", "Astrocyte",
                     "Oligodendrocyte", "OPC", "Neuron", "Neuron", "Microglia",
                     "Neuron", "Neuron", "Astrocyte", "Neuron", "Neuron",
                     "Neuron", "Neuron", "Neuron", "OPC", "Neuron",
                     "Endothelial", "Neuron", "Neuron", "Neuron", "Neuron",
                     "OPC")
names(new.cluster.ids) <- levels(combined) 
combined <- RenameIdents(combined, new.cluster.ids) 
combined@meta.data$summary <- Idents(combined)
```

```{r}
# Detailed Idents
Idents(combined) <- combined@meta.data$seurat_clusters 
new.cluster.ids <- c("Oligodendrocyte_0", "Neuron_1", "Oligodendrocyte_2", "Neuron_3", "Astrocyte_4",
                     "Oligodendrocyte_5", "OPC_6", "Neuron_7", "Neuron_8", "Microglia_9",
                     "Neuron_10", "Neuron_11", "Astrocyte_12", "Neuron_13", "Neuron_14",
                     "Neuron_15", "Neuron_16", "Neuron_17", "OPC_18", "Neuron_19",
                     "Endothelial_20", "Neuron_22", "Neuron_23", "Neuron_24", "Neuron_28",
                     "OPC_30")
names(new.cluster.ids) <- levels(combined) 
combined <- RenameIdents(combined, new.cluster.ids) 
combined@meta.data$detail <- Idents(combined)
```



Source used for cell type assignment
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6329831/
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6671143/ 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6655561/





```{r}
# Visualize data
p1 <- dittoDimPlot(combined, "summary", do.label = TRUE, legend.show = FALSE, labels.size = 3) 
ggsave(filename= paste0(home_dir, "/figures/gse157827/summary.png"), width = 9, height = 7) # name the file as an EPS

p1 <- dittoDimPlot(combined, "detail") 
ggsave(filename= paste0(home_dir, "/figures/gse157827/detail.png"), width = 9, height = 7) # name the file as an EPS

```

```{r}
# Save dataset after clustering
#saveRDS(combined, file = paste0(home_dir, "/sc_human_objects/gse157827_final_filtered.RData"))
combined <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse157827_final_filtered.RData"))
```


```{r}
table(combined@meta.data)
```


# Visualizations for Manuscript

## UMAP

```{r}
# UMAP of Human cells
Idents(combined) <- combined@meta.data$summary
levels(combined) <- c("Astrocyte", "Endothelial", "Microglia", 
                      "Neuron", "Oligodendrocyte", "OPC", "Fibroblast")
p1 <- dittoDimPlot(combined, Idents(combined)) +
  theme(
    axis.title = element_text(size = 16),   # both x and y axis titles
    axis.text  = element_text(size = 14),   # both x and y tick labels
    axis.ticks.length = unit(0.25, "cm")    # physical tick mark length
  )
ggsave(filename = paste0(home_dir, "/Manuscript Figures/GSE157827_umap.pdf"),
       plot = p1, device = cairo_pdf, width = 9, height = 7, units = "in")
```

## DotPlot

```{r}
levels(combined) <- c("Microglia", "Endothelial", "Fibroblast", "Astrocyte", 
                      "OPC", "Oligodendrocyte", "Neuron")
markers <- c("GAD1", "NRGN",
             "ANLN", 
             "VCAN", 
             "CABLES1", "AQP4",
             "DCN",
             "VWF", "FLT1", 
             "CSF1R")
# Dotplot of common human markers
markers <- rev(markers)
p2 <- DotPlot(combined, features = markers, assay = "RNA") + 
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y  = element_text(size = 14),
    axis.title   = element_text(size = 16),
    legend.text  = element_text(size = 12),
    legend.title = element_text(size = 14)
  ) + 
  scale_y_discrete(limits = rev)
ggsave(filename = paste0(home_dir, "/Manuscript Figures/GSE157827_dotplot.pdf"),
       plot = p2, device = cairo_pdf, width = 9, height = 7, units = "in")

```


# Session Information

```{r}
sessionInfo()
```

