---
title: "AD Mouse Processing - GSE208683"
author: "Alexander Bergendorf"
date: "2023-06-15"
output:
  pdf_document: default
  html_document: default
---

# Library Import
```{r}
library(Seurat) 
library(tidyverse) 
library(patchwork) 
library(dittoSeq) 
library(stringr)
library(gprofiler2)
library(harmony)
library(patchwork)
library(cowplot)
library(openxlsx)
library(MAST)
#library(glmGamPoi)

#home_dir <- "C:/Users/aberg/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
home_dir <- "C:/Users/Alexander/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
```

# Normal Object Setup
## Data Import

```{r}
## Reading in mouse data, only normal cells
#rownames <- rep("character", 1)
#control <- rep("integer", 15013)
#other <- rep("NULL", 46936)
#colClasses <- c(rownames, control, other)

#counts <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_UMI_matrix.txt"), sep = "\t", colClasses = colClasses)
#metadata <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_meta_data.txt"), sep = "\t")

#metadata <- metadata[1:15013,]
```


## Create Object

```{r}
#colnames(counts) <- gsub("\\.", "-", colnames(counts))
#wild_type <- CreateSeuratObject(counts = counts, meta.data = metadata)
# These are the previous identifications of the wild type cells from the data source
#table(wild_type@meta.data$SampleID, wild_type@meta.data$CellType)
#rm(counts, metadata)
```


```{r}
#saveRDS(wild_type, file = paste0(home_dir, "/AD_mouse_objects_ortholog/wt_ortholog.RData"))
#wild_type <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/wt_ortholog.RData"))
```


# AD Object Setup
## Data Import
```{r}
## Reading in human data, only 5xFAD cells only
#rownames <- rep("character", 1)
#other1 <- rep("NULL", 31848)
#ad <- rep("integer", 14662)
#other2 <- rep("NULL",15439)
#colClasses <- c(rownames, other1, ad, other2)

#counts <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_UMI_matrix.txt"), sep = "\t", colClasses = colClasses)
#metadata <- read.table(file = paste0(home_dir, "/AD_mouse_raw/GSE208683_Allcell_6m_meta_data.txt"), sep = "\t")

#metadata <- metadata[31849:46510,]
```

## Create Object
```{r}
#colnames(counts) <- gsub("\\.", "-", colnames(counts))
#ad <- CreateSeuratObject(counts = counts, meta.data = metadata)
## These are the previous identifications of the wild type cells from the data source
#table(ad@meta.data$SampleID, ad@meta.data$CellType)
#rm(counts, metadata)
```


```{r}
#saveRDS(ad, file = paste0(home_dir, "/AD_mouse_objects_ortholog/ad_ortholog.RData"))
#ad <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/ad_ortholog.RData"))
```

# Merging and Processing Data

```{r}
ad <- readRDS(file = paste0(home_dir, "/ad_mouse_objects_ortholog/ad_ortholog_gse208683.RData"))
wild_type <- readRDS(file = paste0(home_dir, "/ad_mouse_objects_ortholog/wt_ortholog_gse208683.RData"))

ad <- SplitObject(ad, split.by = "SampleID")
wild_type <- SplitObject(wild_type, split.by = "SampleID")
data_list <- c(ad, wild_type)
#rm(ad, wild_type)
```


```{r}
table(data_list$AD_1@meta.data$orig.ident)
table(data_list$AD_2@meta.data$orig.ident)
```




```{r}
head(data_list$AD_1@meta.data)
# Creating chr vector to loop over for mouse data
names <- c("AD_1", "AD_2", "WT_1", "WT_2")
```

```{r}
mouse_ortholog <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_orthologs_6datasets.RData"))
```


```{r}
# Filter mouse data based on orthologs
for(i in names){
  data_list[[i]] <- data_list[[i]][mouse_ortholog]
}
```



# Calculating Quality Control Metrics


```{r}
for(i in names){ 
  data_list[[i]][["rna.size"]] <- log10(Matrix::colSums(GetAssayData(data_list[[i]], assay = "RNA", layer = "counts")))
  data_list[[i]][["percent.mt"]] <- PercentageFeatureSet(data_list[[i]], pattern = "^mt-")
}
```


```{r}
data_list_test <- merge(x = data_list[[1]], y = data_list[2:length(data_list)], merge.data = TRUE)
df <- data_list_test@meta.data

# RNA size
p1 <- ggplot(df, aes(x = SampleID, y = rna.size, fill = SampleID)) +
  geom_violin() +
  rasterise(geom_jitter(width = 0.25, size = 0.1, alpha = 0.4), dpi = 300) +
  theme_classic(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Sample", y = "RNA Size")

ggsave(paste0(home_dir, "/Manuscript Figures/New Data/rnasize_qc_gse208683.pdf"),
       plot = p1, device = cairo_pdf, width = 6, height = 5, units = "in")

# Percent MT
p2 <- ggplot(df, aes(x = SampleID, y = percent.mt, fill = SampleID)) +
  geom_violin() +
  rasterise(geom_jitter(width = 0.25, size = 0.1, alpha = 0.4), dpi = 300) +
  theme_classic(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  labs(x = "Sample", y = "Percent MT")

ggsave(paste0(home_dir, "/Manuscript Figures/New Data/mt_qc_gse208683.pdf"),
       plot = p2, device = cairo_pdf, width = 6, height = 5, units = "in")
```


```{r}
rna_size_min <- list() 
rna_size_max <- list() 

# Calculates 3 MAD from the median rna-size, filters out if outside that range
for(i in names){ 
  rna_size_min[[i]] <- median(data_list[[i]]@meta.data$rna.size) - (3*mad(data_list[[i]]@meta.data$rna.size)) 
  
  rna_size_max[[i]] <- median(data_list[[i]]@meta.data$rna.size) + (3*mad(data_list[[i]]@meta.data$rna.size)) 
} 
```


```{r}
# Save QC metric DF
qc_metrics <- c() 
for(i in names){ 
  preqc_cells <- ncol(data_list[[i]]) 
  qc_metrics <- rbind(qc_metrics, preqc_cells) 
}

colnames(qc_metrics) <- "preqc_cells" 
row.names(qc_metrics) <- NULL 
preqc_metrics <- as.data.frame(qc_metrics)
```


```{r}
# Filter based on QC metrics
for(i in names){ 
  temp_max <- rna_size_max[[i]] 
  temp_min <-  rna_size_min[[i]] 
  data_list[[i]] <- subset(data_list[[i]], subset = percent.mt < 10 & rna.size < temp_max & rna.size > temp_min) 
  
}
```


```{r}
# Finish QC Table
qc_metrics <- c() 
for(i in names){ 
  preqc_cells <- ncol(data_list[[i]]) 
  qc_metrics <- rbind(qc_metrics, preqc_cells) 
}
```

```{r}
colnames(qc_metrics) <- "postqc_cells" 
row.names(qc_metrics) <- NULL 
postqc_metrics <- as.data.frame(qc_metrics)
cellcounts <- cbind(preqc_metrics, postqc_metrics)
cellcounts
```

```{r}
write.csv(cellcounts, file = paste0(home_dir, "/Manuscript and Figures/New Data/mouse_cellcounts_ortholog_gse208683.csv"))
```


# Check Cell Cycle after clustering

```{r}
# S phase genes
mmus_s <- gorth(cc.genes.updated.2019$s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
# G2M phase genes
mmus_g2m <- gorth(cc.genes.updated.2019$g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name

# Normalizes data and calculates cell-cycle scoring to be regressed
for(i in names){ 
    data_list[[i]] <- NormalizeData(data_list[[i]]) 
    data_list[[i]] <- CellCycleScoring(data_list[[i]], s.features = mmus_s, g2m.features = mmus_g2m) 
}
```

# SCTransform

```{r}
gc()
#Applies SCTransform individually to each sample 
data_list <- lapply(X = data_list, FUN = SCTransform,  method ="glmGamPoi", return.only.var.genes = FALSE, vars.to.regress = c("S.Score", "G2M.Score"))
gc()
```


```{r}
# Save data after SCTransform, computationally intensive
#saveRDS(data_list, file = paste0(home_dir, "/AD_mouse_objects_ortholog/data_list_ortholog_gse208683.RData"))
#data_list <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/data_list_ortholog.RData"))
```



```{r}
# Calculates variable features across samples for integration
variable.features <- SelectIntegrationFeatures(object.list = data_list, nfeatures = 3000)
```

# Batch Correction

```{r}
# Merges samples
data_list <- merge(x = data_list[[1]], y = data_list[2:length(data_list)], merge.data=TRUE) 

# Assigns variable features calculated across samples
VariableFeatures(data_list) <- variable.features 
gc()
```


```{r}
# Performs PCA
data_list <- RunPCA(data_list, verbose = FALSE) 
```

```{r}
variance <- (data_list@reductions$pca@stdev)^2
variance <- variance / sum(variance)
variance
```



```{r}
# PC Plot before integration occurs
options(repr.plot.height = 12, repr.plot.width = 20) 
p1 <- DimPlot(object = data_list, reduction = "pca", pt.size = 2, group.by = "SampleID", raster = TRUE) 
#p2 <- VlnPlot(object = data_list, features = "PC_1", group.by = "orig.ident", pt.size = 0) 
#plot_grid(p1,p2)
ggsave(filename= paste0(home_dir, "/Manuscript Figures/New Data/gse208683_pre_batch.pdf"),
       plot = p1, device = cairo_pdf, width = 6, height = 5, units = "in") # name the file as an EPS
```

We should batch correct.

```{r}
# Running Harmony  
options(repr.plot.height = 8, repr.plot.width = 16) 

data_list <- RunHarmony(data_list, assay.use = "SCT", group.by.vars = "SampleID", plot_convergence = TRUE) 
ggsave(paste0(home_dir, "/Manuscript and Figures/New Data/convergenece_plot_gse208683.pdf"))
```

```{r}
variance <- (data_list@reductions$harmony@stdev)^2
variance <- variance / sum(variance)
variance
```


```{r}
# PC plot after integration occurs
p1 <- DimPlot(object = data_list, reduction = "harmony", pt.size = 2, group.by = "SampleID", raster = TRUE) 

#p2 <- VlnPlot(object = data_list, features = "harmony_1", group.by = "orig.ident", pt.size = 0) 
ggsave(filename= paste0(home_dir, "/Manuscript Figures/New Data/gse208683_post_batch.pdf"),
       plot = p1, device = cairo_pdf, width = 6, height = 5, units = "in") # name the file as an EPS
```



```{r}
# Elbowplot to select PCs for clustering
ElbowPlot(data_list, reduction = "harmony", n = 50) 
```

Start with 30 PCS, check separation


# Reducing Dimensionality and Clustering

```{r}
#Performing Dimensional Reduction 

# UMAP
data_list <- RunUMAP(data_list, reduction = "harmony", dims = 1:30, verbose = FALSE) 

# Find Neighboring cells 
data_list <- FindNeighbors(data_list, reduction = "harmony", dims = 1:30, verbose = FALSE) 

# knn clustering
data_list <- FindClusters(data_list, resolution = 1, verbose = FALSE) 
```



```{r}
# Visualize data
dittoDimPlot(data_list, "seurat_clusters", do.label = TRUE, legend.show = FALSE, labels.size = 3) 
dittoDimPlot(data_list, "percent.mt") 

```

```{r}
# Save dataset after clustering
#saveRDS(data_list, file = paste0(home_dir, "/AD_mouse_objects_ortholog/gse208683_final.RData"))
data_list <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/gse208683_final.RData"))
```




# Visualizing Data


```{r}
# Visualization of some markers
marker_list <- c("Mrc1", "Trem2", "Igfbp7", "Col1a1", "Aldh1l1", "Hes5", "Pdgfra", "Matn4",
                 "Mog", "Mobp", "Mbp", "Satb2", "Reln")
Idents(data_list) <- data_list@meta.data$seurat_clusters
#for(i in marker_list){
  #p1 <- dittoDimPlot(data_list, var = i, assay = "RNA")
  #ggsave(filename = paste0(home_dir, "/figures/gse208683/markers/umap_", i, ".png"),
        # plot = p1, width = 9, height = 7)
#}

# Dot plot of all markers across clusters
p2 <- DotPlot(data_list, features = marker_list, assay = "RNA") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_y_discrete(limits = rev)
ggsave(filename = paste0(home_dir, "/figures/gse208683/markers/dotplot_markers.png"),
       plot = p2, width = 9, height = 7)
p2
```

# Finding Cluster Identifiers

```{r}
# Get unique identifiers
unique_idents <- as.character(unique(data_list@meta.data$seurat_clusters))
unique_idents

# Initialize an empty list to store the results
markers_list <- list()

# Loop over each identifier and run FindMarkers
for (id in unique_idents) {
  markers_list[[id]] <- FindMarkers(
    object = data_list,
    ident.1 = id,
    only.pos = TRUE,
    assay = "RNA",
    max.cells.per.ident = 300
  )
}

saveRDS(markers_list, paste0(home_dir, "/all_cell_markers_mouse_ortholog_gse208683.RData"))

```

```{r}
markers_list <- readRDS(paste0(home_dir, "/all_cell_markers_mouse_ortholog_gse208683.RData"))
# Get counts per cluster from the Seurat object
cluster_counts <- table(data_list$seurat_clusters)

# Rename each list element: "clusternumber_clustercount"
for (cl in names(markers_list)) {
  names(markers_list)[names(markers_list) == cl] <- paste0(cl, "_n", cluster_counts[[cl]])
}
```



```{r}
write.xlsx(markers_list, file = paste0(home_dir, "/all_cell_markers_mouse_ortholog_gse208683.xlsx"), rowNames = TRUE)
```

0- Oligodendrocyte
1- Neuron
2- Neuron
3- Astrocyte
4- Neuron
5- Neuron
6- Neuron
7- Neuron
8- Neuron
9- Neuron
10- Neuron
11- Microglia
12- OPC
13- Neuron
14- Neuron
15- Endothelial
16- Astrocyte
17- Neuron
18- Neuron
19- Doublet
20- Neuron
21- Neuron
22- Neuron
23- Neuron
24- Neuron
25- Astrocyte
26- Neuron
27- Neuron
28- Neuron
29- Endothelial
30- Microglia
31- Doublet
32- Doublet
33- OPC



# Assign Idents

```{r}
data_list <- subset(data_list, subset = SCT_snn_res.1 %in% c(19, 31, 32), invert = TRUE)
# Summary Idents
Idents(data_list) <- data_list@meta.data$seurat_clusters
new.cluster.ids <- c(
  "Oligodendrocyte",           # 0
  "Neuron",                    # 1
  "Neuron",                    # 2
  "Astrocyte",                 # 3
  "Neuron",                    # 4
  "Neuron",                    # 5
  "Neuron",                    # 6
  "Neuron",                    # 7
  "Neuron",                    # 8
  "Neuron",                    # 9
  "Neuron",                    # 10
  "Microglia",                 # 11
  "Oligodendrocyte Precursor", # 12
  "Neuron",                    # 13
  "Neuron",                    # 14
  "Endothelial",               # 15
  "Astrocyte",                 # 16
  "Neuron",                    # 17
  "Neuron",                    # 18
  "Neuron",                    # 20
  "Neuron",                    # 21
  "Neuron",                    # 22
  "Neuron",                    # 23
  "Neuron",                    # 24
  "Astrocyte",                 # 25
  "Neuron",                    # 26
  "Neuron",                    # 27
  "Neuron",                    # 28
  "Endothelial",               # 29
  "Microglia",                 # 30
  "Oligodendrocyte Precursor"  # 33
)
names(new.cluster.ids) <- levels(data_list)
data_list <- RenameIdents(data_list, new.cluster.ids)
data_list@meta.data$summary <- Idents(data_list)
```

```{r}
# Detailed Idents
Idents(data_list) <- data_list@meta.data$seurat_clusters
new.cluster.ids <- c(
  "Oligodendrocyte_0", "Neuron_1", "Neuron_2", "Astrocyte_3", "Neuron_4",
  "Neuron_5", "Neuron_6", "Neuron_7", "Neuron_8", "Neuron_9",
  "Neuron_10", "Microglia_11", "OPC_12", "Neuron_13", "Neuron_14",
  "Endothelial_15", "Astrocyte_16", "Neuron_17", "Neuron_18", "Neuron_20",
  "Neuron_21", "Neuron_22", "Neuron_23", "Neuron_24", "Astrocyte_25",
  "Neuron_26", "Neuron_27", "Neuron_28", "Endothelial_29", "Microglia_30",
  "OPC_33"
)
names(new.cluster.ids) <- levels(data_list)
data_list <- RenameIdents(data_list, new.cluster.ids)
data_list@meta.data$detail <- Idents(data_list)
```



Source used for cell type assignment
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6329831/
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6671143/ 
https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6655561/


```{r}
# Save seurat object with cell types identified
#saveRDS(data_list, file = paste0(home_dir, "/AD_mouse_objects_ortholog/gse208683_final_filtered.RData"))
data_list <- readRDS(file = paste0(home_dir, "/AD_mouse_objects_ortholog/gse208683_final_filtered.RData"))
```

```{r}
table(data_list@meta.data$orig.ident, data_list@meta.data$SampleID)
```


```{r}
table(data_list@meta.data$seurat_clusters, data_list@meta.data$summary)
table(data_list@meta.data$seurat_clusters, data_list@meta.data$Genotype)

```

11 and 30

```{r}
data_list <- subset(data_list, subset = summary == "Microglia")
saveRDS(data_list, file = paste0(home_dir, "/AD_mouse_objects_ortholog/gse208683_microglia_pre.RData"))
```


```{r}
table(data_list@meta.data$orig.ident, data_list@meta.data$Genotype)
```


# Visualizations for Manuscript

## UMAP

```{r}
data_list@meta.data$summary <- if_else(data_list@meta.data$summary == "Oligodendrocyte Precursor", "OPC", data_list@meta.data$summary)
```



```{r}
# UMAP of Human cells
Idents(data_list) <- data_list@meta.data$summary
levels(data_list) <- c("Astrocyte", "Endothelial", "Microglia", 
                      "Neuron", "Oligodendrocyte", "OPC")
p1 <- dittoDimPlot(data_list, Idents(data_list)) +
  theme(
    axis.title = element_text(size = 16),   # both x and y axis titles
    axis.text  = element_text(size = 14),   # both x and y tick labels
    axis.ticks.length = unit(0.25, "cm")    # physical tick mark length
  )
ggsave(filename = paste0(home_dir, "/figures/manuscript/GSE208683_umap.pdf"),
       plot = p1, device = cairo_pdf, width = 9, height = 7, units = "in")
p1
```

## DotPlot

```{r}
levels(data_list) <- c("Microglia", "Endothelial", "Astrocyte", 
                      "OPC", "Oligodendrocyte", "Neuron")

markers <- c("Mrc1", "Trem2", "Igfbp7", "Aldh1l1", "Hes5", "Pdgfra", "Matn4",
                 "Mog", "Mobp", "Mbp", "Satb2", "Reln")

# Dotplot of common human markers
p2 <- DotPlot(data_list, features = markers, assay = "RNA") + 
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 14),
    axis.text.y  = element_text(size = 14),
    axis.title   = element_text(size = 16),
    legend.text  = element_text(size = 12),
    legend.title = element_text(size = 14)
  ) + 
  scale_y_discrete(limits = rev)
ggsave(filename = paste0(home_dir, "/figures/manuscript/GSE208683_dotplot.pdf"),
       plot = p2, device = cairo_pdf, width = 9, height = 7, units = "in")
p2

```

# Session Information

```{r}
sessionInfo()
```