---
title: "AD TransComp-R PCA"
author: "aberg Bergendorf"
date: "2023-08-07"
output: html_document
---


# Library Import

```{r}
library(Seurat)
library(mixOmics)
library(tidyverse)
library(orthogene)
library(dittoSeq)
library(openxlsx)
#library(PCAtools)
library(MASS)
library(glmnet)
library(pROC)

home_dir <- "C:/Users/aberg/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
#home_dir <- "C:/Users/Alexander/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
```


# Loading Models for PC Construction

```{r}
mouse_gse208863 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))
```

```{r}
merged <- merge(mouse_gse208863, mouse_gse140510, add.cell.ids = c("gse208863", "gse140510"), project = "merged")

# In Seurat v5, layers are split by default after merge. Join them so the assay is truly merged:
merged <- JoinLayers(merged)
```




## Run PCA 

```{r}
# Rerunning PCA after filtering
DefaultAssay(merged) <- "RNA"
merged <- NormalizeData(merged)
merged <- FindVariableFeatures(merged, selection.method = "vst", nfeatures = 3250)
all.genes <- rownames(merged)
merged <- ScaleData(merged, features = all.genes)

# Manual PCA via irlba (Seurat's RunPCA silently drops ~500 features in v5)
sd_mat <- GetAssayData(merged, layer = "scale.data")[VariableFeatures(merged), ]
pca_irlba <- irlba::irlba(A = t(sd_mat), nv = 50)

embeddings <- pca_irlba$u %*% diag(pca_irlba$d)
loadings   <- pca_irlba$v
stdev      <- pca_irlba$d / sqrt(max(1, ncol(sd_mat) - 1))

rownames(embeddings) <- colnames(sd_mat)
colnames(embeddings) <- paste0("PC_", 1:50)
rownames(loadings)   <- rownames(sd_mat)
colnames(loadings)   <- paste0("PC_", 1:50)

merged[["pca"]] <- CreateDimReducObject(
  embeddings = embeddings,
  loadings   = loadings,
  stdev      = stdev,
  key        = "PC_",
  assay      = "RNA"
)

# Sanity check
dim(Loadings(merged, "pca"))    # 3250 × 50
dim(Embeddings(merged, "pca"))  # ncells × 50
```




```{r}
# Variances of mouse PCs
stdev <- merged@reductions$pca@stdev
variance <- stdev^2
variance <- variance / sum(variance)
variance
write.csv(variance, paste0(home_dir, "/pca_variances_merged.csv"))
```

```{r}
variance <- read.csv(paste0(home_dir, "/pca_variances_merged.csv"))
colnames(variance) <- c("PC", "sdev")
variance$cum_var <- cumsum(variance$sdev)
variance
```

## Cumulative Variance Plot

```{r}
#svg('C:/Users/aberg/OneDrive/Desktop/AD Microglia Manuscript Figures/Mouse_variance_explained_pca.svg', width = 5, height = 4)
#ggplot(data = variance, aes(x = PC, y = cum_var, group = 1)) +
  #geom_line(color = "black") +  # Changed line color to blue
  #geom_point(color = "black") +  # Added points with the same color
  #geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +  # Added red dashed line at 0.80 cum_var
  #geom_point(data = subset(variance, cum_var == 0.80), aes(x = PC, y = cum_var), color = "red", size = 3) +  # Add red point at the threshold
  #ylab("Cumulative Variance") +# Changed y-axis label
  #theme_classic() + theme_bw()
#dev.off()
```


## UMAP of Cells

```{r}
# Cluster cells and visualize on UMAP for paper visualization
merged <- FindNeighbors(merged, dims = 1:5)
merged <- FindClusters(merged, resolution = 0.5)
merged <- RunUMAP(merged, dims = 1:5)
```


```{r}
p1 <- dittoDimPlot(merged, "phenotype", color.panel = c("#E69F00", "#56B4E9"))
p1
ggsave(filename= paste0(home_dir, "/figures/manuscript/umap_phenotype_merged_mouse.pdf"), width = 9, height = 7) # name the file as an EPS
```


```{r}
# Counts of mouse microglial cells
count_table <- table(merged@meta.data$phenotype)
count_table
write.csv(count_table, file = paste0(home_dir, "/phenotype_counts_merged.csv"))
```

```{r}
# Counts of mouse microglial cells
sample_table <- table(merged@meta.data$orig.ident)
sample_table
write.csv(sample_table, file = paste0(home_dir, "/sample_counts_merged.csv"))
```


```{r}
p1 <- ElbowPlot(merged, ndims = 50)
p1
ggsave(filename= paste0(home_dir, "/figures/manuscript/elbowplot_clustering2_merged.pdf"), width = 9, height = 7) # name the file as an EPS
```



## Loading mouse principal components

```{r}
#saveRDS(merged, paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_pca_loadings_merged.RData"))
merged <- readRDS(paste0(home_dir, "/AD_mouse_objects_ortholog/mouse_pca_loadings_merged.RData"))
```

```{r}
# Pull principal components from Seurat object
principal_components <- merged@reductions$pca@feature.loadings
principal_components <- as.data.frame(principal_components)
```

```{r}
saveRDS(principal_components, paste0(home_dir, "/mouse_feature_loadings_pca_merged.RData"))
```


```{r include=FALSE}
# plot pc projections to visualize
for(i in 2:50){
p1 <- DimPlot(merged, reduction = "pca", dims = c(1, i), group.by = "phenotype")
ggsave(filename = paste0(home_dir, "/pc_plots/pc_plot_pca_merged_on_merged/pc_", i, ".pdf"))
}
```

```{r}
p1 <- dittoDimPlot(merged, reduction.use = "pca", dim.1 = 2, dim.2 = 5, var = "phenotype")
ggsave(filename = paste0(home_dir, "/figures/manuscript/pc_2_5_mouse.pdf"), plot = p1)
```


## Checking which PCs are significant in mice


```{r}
merged <- AddMetaData(merged, merged@reductions$pca@cell.embeddings)
```

```{r}
#summary(ad.model)
merged@meta.data$study <- if_else(merged@meta.data$orig.ident %in% c("1_HS", "2_HS", "5_HS", "6_HS"), "GSE208683", "GSE140510")
```


```{r}
pcs <- paste0("PC_", 1:36)

# Test each PC against study (linear model; use Kruskal-Wallis if non-normal)
study_pvals <- sapply(pcs, function(pc) {
  f <- as.formula(paste(pc, "~ study"))
  anova(lm(f, data = merged@meta.data))$"Pr(>F)"[1]
})

# Adjust for multiple testing
study_padj <- p.adjust(study_pvals, method = "BH")

results <- data.frame(
  PC = pcs,
  p_value = study_pvals,
  p_adj = study_padj,
  study_specific = study_padj < 0.05
)

# PCs NOT significantly associated with study (safe to use)
not_study_specific <- results$PC[!results$study_specific]
print(results)
print(not_study_specific)
```

```{r}
as.matrix(results)
```


The PCs that are separative of study-specific biology include PC_1, PC_3, and PC_6 

```{r}
merged$dv <- if_else(merged$phenotype == "C", 0, 1)
```

```{r}
ad.model <- glm(dv ~ PC_2 + PC_4 + PC_5 + 
                  PC_7 + PC_8 + PC_9 + PC_10 +
                  PC_11 + PC_12 + PC_13 + PC_14 + PC_15 +
                  PC_16 + PC_17 + PC_18 + PC_19 + PC_20 +
                  PC_21 + PC_22 + PC_23 + PC_24 + PC_25 +
                  PC_26 + PC_27 + PC_28 + PC_29 + PC_30 +
                  PC_31 + PC_32 + PC_33
                  , data = merged@meta.data, family = binomial)
```

```{r message=FALSE, include=FALSE}
final_model <- stepAIC(ad.model, direction = "both")
```

```{r}
summary(final_model)
```






# GSE175814 (initial human)
## Run PCA  for visualization

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse175814_microglia_final.RData"))
```



```{r}
# Rerunning PCA again
DefaultAssay(human_gse175814) <- "RNA"
human_gse175814 <- NormalizeData(human_gse175814)
all.genes <- rownames(human_gse175814)
human_gse175814 <- ScaleData(human_gse175814, features = all.genes)
human_gex_matrix <- as.data.frame(LayerData(human_gse175814, assay = "RNA", layer = "scale.data"))
human_gse175814 <- RunPCA(human_gse175814, features = VariableFeatures(object = human_gse175814))
```



## Plotting Human GEX Matrix

```{r}
# Plotting on UMAP for visualization
human_gse175814 <- FindNeighbors(human_gse175814, dims = 1:5)
human_gse175814 <- FindClusters(human_gse175814, resolution = 0.5)
human_gse175814 <- RunUMAP(human_gse175814, dims = 1:5)
```

```{r}
p2 <- dittoDimPlot(human_gse175814, "phenotype", color.panel = c("#F8766D", "#00BFC4"))
p2
ggsave(filename= paste0(home_dir, "/figures/manuscript/umap_phenotype_gse175814.pdf"), width = 9, height = 7) # name the file as an EPS

```

```{r}
# Count of human microglial cells
count_table <- table(human_gse175814@meta.data$phenotype)
count_table
write.csv(count_table, file = paste0(home_dir, "/phenotype_counts_gse175814.csv"))
```

```{r}
# Count of human microglial cells
count_table <- table(human_gse175814@meta.data$orig.ident)
count_table
write.csv(count_table, file = paste0(home_dir, "/sample_counts_gse175814.csv"))
```






```{r}
# Confirming only orthologs in GEX matrix
ortholog_mat <- convert_orthologs(gene_df = row.names(human_gex_matrix),
                                    input_species = "human",
                                    output_species = "mouse",
                                    non121_strategy = "drop_both_species")

# Filtering the orthologs based on the 3250 principal component loadings
ortholog_mat <- ortholog_mat %>%
  filter(row.names(ortholog_mat) %in% row.names(principal_components))
        

# Filtering the human gex matrix based on the 3250 principal component loadings
human_gex_matrix <- human_gex_matrix %>%
  filter(row.names(human_gex_matrix) %in% ortholog_mat$input_gene)

sum(row.names(human_gex_matrix) == ortholog_mat$input_gene)

# Assigning ortholog names to human gene expression matrix
row.names(human_gex_matrix) <- row.names(ortholog_mat)

# Arrange alphabetically 
principal_components  <- principal_components %>% arrange(row.names(principal_components))

# Arrange alphabetically
human_gex_matrix <- human_gex_matrix %>%
  arrange(row.names(human_gex_matrix))

```

Now the human data is put in mouse space

```{r}
sum(row.names(human_gex_matrix) == row.names(principal_components))
```



```{r}
# converting to matrix
principal_components <- as.matrix(principal_components)
human_gex_matrix <- as.matrix(human_gex_matrix)

# Ensuring equal rows, equals 3250
ncol(t(principal_components))
nrow(human_gex_matrix)

# Multiplies the transpose of the principal component matrix by the human gex matrix
multiplication_2 <- t(human_gex_matrix) %*% principal_components

# Creates new dimensional reduction on human Seurat object with the pc projections
human_gse175814[["pca_projections"]] <- CreateDimReducObject(embeddings = multiplication_2, key = "pca_proj")
```

```{r}
human_gse175814$dv <- if_else(human_gse175814$phenotype == "C", 0, 1)
```

```{r}
human_gse175814 <- AddMetaData(human_gse175814, human_gse175814@reductions$pca_projections@cell.embeddings)
```


```{r}
saveRDS(human_gse175814, paste0(home_dir, "/sc_human_objects/merged_pca_on_gse175814_microglia_final_pca.RData"))
```



## Running GLMs on Projected GSE175814
### 80% variance

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse175814_microglia_final_pca.RData"))
```




```{r}
# Run GLM
ad.model <- glm(dv ~ pcaproj_2 + pcaproj_4 + pcaproj_5 + 
                  pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 
                  , data = human_gse175814@meta.data, family = binomial)

```

```{r}
summary(ad.model)
```

```{r}
saveRDS(ad.model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse175814_glm_80var.RData"))
```



### AIC Optimization

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse175814_microglia_final_pca.RData"))
```

```{r}
# Run GLM
ad.model <- glm(dv ~  pcaproj_2 +  pcaproj_4 + pcaproj_5 + 
                   pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33
                  , data = human_gse175814@meta.data, family = binomial)
```

```{r message=FALSE, include=FALSE}
final_model <- stepAIC(ad.model, direction = "both")
```

```{r}
summary(final_model)
```

```{r}
saveRDS(final_model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse175814_glm_aic_optimized.RData"))
```

### Mixed Model

```{r}
library(lme4)
library(lmerTest)  # for p-values

meta <- human_gse175814@meta.data
meta$dv <- factor(meta$dv)
meta$orig.ident <- factor(meta$orig.ident)

# Mixed-effects logistic regression with donor as random intercept
ad.model.mixed <- glmer(
  dv ~ pcaproj_2 + pcaproj_4 + pcaproj_5 + 
       pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
       pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
       pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
       pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
       pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
       pcaproj_31 + pcaproj_32 + pcaproj_33 +
       (1 | orig.ident),
  data = meta,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(ad.model.mixed)
```




### PC Plots

```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse175814_microglia_final_pca.RData"))
```





```{r include=FALSE}
# plot pc projections to visualize
for(i in 2:50){
p1 <- DimPlot(human_gse175814, reduction = "pca_projections", dims = c(1, i), group.by = "phenotype")
ggsave(filename = paste0(home_dir, "/pc_plots/pc_plot_pca_merged_on_gse175814/pc_", i, ".pdf"))
}
```


```{r}
p1 <- dittoDimPlot(human_gse175814, reduction.use = "pca_projections", dim.1 = 2, dim.2 = 5, var = "phenotype")
ggsave(filename = paste0(home_dir, "/figures/manuscript/pc_2_5_gse175814.pdf"), plot = p1)
```


```{r}
dittoDimPlot(
  object = human,                   # Your Seurat object
  reduction.use = "spca_projections", # Specify the reduction method
  dim.1 = 3,                        # Specify the first dimension
  dim.2 = 5,                        # Specify the second dimension
  var = "Phase"                  # Metadata variable for coloring
)

dittoPlot(
  object = human,             # Your Seurat object
  var = "G2M.Score",          # The y-axis variable
  group.by = "Genotype"       # Specify that you want a violin plot
)
```

# GSE157827 (new human)
## Run PCA  for visualization

```{r}
human_gse157827 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse157827_microglia_final.RData"))
```



```{r}
#human_gse157827 <- subset(human_gse157827, subset = orig.ident %in% c("AD5", "AD10", "AD19",
                                                                       #"NC3", "NC7", "NC11", #"NC12", "NC14", "NC15", "NC16", "NC17", "NC18"))
```

```{r}
table(human_gse157827@meta.data$phenotype)
```



```{r}
# Rerunning PCA again
DefaultAssay(human_gse157827) <- "RNA"
human_gse157827 <- NormalizeData(human_gse157827)
all.genes <- rownames(human_gse157827)
human_gse157827 <- ScaleData(human_gse157827, features = all.genes)
human_gex_matrix <- as.data.frame(LayerData(human_gse157827, assay = "RNA", layer = "scale.data"))
human_gse157827 <- RunPCA(human_gse157827, features = VariableFeatures(object = human_gse157827))
```



## Plotting Human GEX Matrix

```{r}
# Plotting on UMAP for visualization
human_gse157827 <- FindNeighbors(human_gse157827, dims = 1:5)
human_gse157827 <- FindClusters(human_gse157827, resolution = 0.5)
human_gse157827 <- RunUMAP(human_gse157827, dims = 1:5)
```

```{r}
p2 <- dittoDimPlot(human_gse157827, "phenotype", color.panel = c("#F8766D", "#00BFC4"))
p2
ggsave(filename= paste0(home_dir, "/figures/manuscript/umap_phenotype_human_gse157827.pdf"), width = 9, height = 7) # name the file as an EPS

```

```{r}
# Count of human microglial cells
count_table <- table(human_gse157827@meta.data$phenotype)
count_table
write.csv(count_table, file = paste0(home_dir, "/phenotype_counts_gse157827.csv"))
```



```{r}
# Count of human microglial cells
count_table <- table(human_gse157827@meta.data$orig.ident)
count_table
write.csv(count_table, file = paste0(home_dir, "/sample_counts_gse157827.csv"))
```







```{r}
# Loading pre-computed 1:1 ortholog matrix (replaces convert_orthologs() call)
# Row names = human gene symbols, input_gene column = mouse ortholog
ortholog_mat <- read.csv(paste0(home_dir, "/ortholog_matrix_IMPORTANT_6datasets.csv"),
                         row.names = 1,
                         stringsAsFactors = FALSE)

# Restricting to orthologs whose human symbol is present in the human GEX matrix
# (mirrors the implicit filter that convert_orthologs applied to row.names(human_gex_matrix))
ortholog_mat <- ortholog_mat %>%
  filter(row.names(ortholog_mat) %in% row.names(human_gex_matrix))

# Filtering the orthologs based on the 3250 principal component loadings
ortholog_mat <- ortholog_mat %>%
  filter(input_gene %in% row.names(principal_components))

# Filtering the human gex matrix based on the 3250 principal component loadings
# NOTE: filter on row.names(human_gex_matrix), which are HUMAN symbols == row.names(ortholog_mat)
human_gex_matrix <- human_gex_matrix %>%
  filter(row.names(human_gex_matrix) %in% row.names(ortholog_mat))

# Align ordering so row i of human_gex_matrix corresponds to row i of ortholog_mat
ortholog_mat     <- ortholog_mat[order(row.names(ortholog_mat)), , drop = FALSE]
human_gex_matrix <- human_gex_matrix[order(row.names(human_gex_matrix)), , drop = FALSE]

# Sanity check before renaming
stopifnot(all(row.names(human_gex_matrix) == row.names(ortholog_mat)))

# Assigning ortholog (mouse) names to human gene expression matrix
row.names(human_gex_matrix) <- ortholog_mat$input_gene

# Arrange alphabetically (now both keyed on mouse symbols)
principal_components <- principal_components %>% arrange(row.names(principal_components))
human_gex_matrix     <- human_gex_matrix     %>% arrange(row.names(human_gex_matrix))

# Final sanity check that row names align before any matrix multiplication downstream
stopifnot(all(row.names(human_gex_matrix) == row.names(principal_components)))
```

Now the human data is put in mouse space

```{r}
sum(row.names(human_gex_matrix) == row.names(principal_components))
```



```{r}
# converting to matrix
principal_components <- as.matrix(principal_components)
human_gex_matrix <- as.matrix(human_gex_matrix)

# Ensuring equal rows, equals 3250
ncol(t(principal_components))
nrow(human_gex_matrix)

# Multiplies the transpose of the principal component matrix by the human gex matrix
multiplication_2 <- t(human_gex_matrix) %*% principal_components

# Creates new dimensional reduction on human Seurat object with the pc projections
human_gse157827[["pca_projections"]] <- CreateDimReducObject(embeddings = multiplication_2, key = "pca_proj")
```

```{r}
human_gse157827$dv <- if_else(human_gse157827$phenotype == "C", 0, 1)
```

```{r}
human_gse157827 <- AddMetaData(human_gse157827, human_gse157827@reductions$pca_projections@cell.embeddings)
```


```{r}
saveRDS(human_gse157827, paste0(home_dir, "/sc_human_objects/merged_pca_on_gse157827_microglia_final_pca.RData"))
```



## Running GLMs on Projected Data
### 80% variance

```{r}
human_gse157827 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse157827_microglia_final_pca.RData"))
```

```{r}
table(human_gse157827@meta.data$phenotype, table(human_gse))
```


```{r}
# Run GLM
ad.model <- glm(dv ~  pcaproj_2 + pcaproj_4 + pcaproj_5 + 
                  pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 
                  , data = human_gse157827@meta.data, family = binomial)

```

```{r}
summary(ad.model)
```

```{r}
saveRDS(ad.model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse157827_glm_80var.RData"))
```


### AIC Optimization

```{r}
human_gse157827 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse157827_microglia_final_pca.RData"))
```

```{r}
# Run GLM
ad.model <- glm(dv ~ pcaproj_2 + pcaproj_4 + pcaproj_5 + 
                  pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 
                  , data = human_gse157827@meta.data, family = binomial)
```

```{r message=FALSE, include=FALSE}
final_model <- stepAIC(ad.model, direction = "both")
```

```{r}
summary(final_model)
```

```{r}
saveRDS(final_model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse157827_glm_aic_optimized.RData"))
```


### PC Plots

```{r}
human_gse157827 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse157827_microglia_final_pca.RData"))
```



```{r include=FALSE}
# plot pc projections to visualize
for(i in 2:50){
p1 <- DimPlot(human_gse157827, reduction = "pca_projections", dims = c(1, i), group.by = "phenotype")
ggsave(filename = paste0(home_dir, "/pc_plots/pc_plot_pca_merged_on_gse157827/pc_", i, ".pdf"))
}
```

```{r}
p1 <- dittoDimPlot(human_gse157827, reduction.use = "pca_projections", dim.1 = 2, dim.2 = 5, var = "phenotype")
ggsave(filename = paste0(home_dir, "/figures/manuscript/pc_2_5_gse157827.pdf"), plot = p1)
```


```{r}
dittoDimPlot(
  object = human,                   # Your Seurat object
  reduction.use = "spca_projections", # Specify the reduction method
  dim.1 = 3,                        # Specify the first dimension
  dim.2 = 5,                        # Specify the second dimension
  var = "Phase"                  # Metadata variable for coloring
)

dittoPlot(
  object = human,             # Your Seurat object
  var = "G2M.Score",          # The y-axis variable
  group.by = "Genotype"       # Specify that you want a violin plot
)
```


# GSE174367 (new human)
## Run PCA  for visualization

```{r}
human_gse174367 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse174367_microglia_final.RData"))
```

```{r}
table(human_gse174367@meta.data$SampleID, human_gse174367@meta.data$Diagnosis)
```


```{r}
# Rerunning PCA again
DefaultAssay(human_gse174367) <- "RNA"
human_gse174367 <- NormalizeData(human_gse174367)
all.genes <- rownames(human_gse174367)
human_gse174367 <- ScaleData(human_gse174367, features = all.genes)
human_gex_matrix <- as.data.frame(LayerData(human_gse174367, assay = "RNA", layer = "scale.data"))
human_gse174367 <- RunPCA(human_gse174367, features = VariableFeatures(object = human_gse174367))
```



## Plotting Human GEX Matrix

```{r}
# Plotting on UMAP for visualization
human_gse174367 <- FindNeighbors(human_gse174367, dims = 1:5)
human_gse174367 <- FindClusters(human_gse174367, resolution = 0.5)
human_gse174367 <- RunUMAP(human_gse174367, dims = 1:5)
```



```{r}
p2 <- dittoDimPlot(human_gse174367, "Diagnosis", color.panel = c("#F8766D", "#00BFC4"))
p2
ggsave(filename= paste0(home_dir, "/figures/manuscript/umap_phenotype_human_gse174367.pdf"), width = 9, height = 7) # name the file as an EPS

```

```{r}
# Count of human microglial cells
count_table <- table(human_gse174367@meta.data$Diagnosis)
count_table
write.csv(count_table, file = paste0(home_dir, "/phenotype_counts_gse174367.csv"))
```



```{r}
# Count of human microglial cells
count_table <- table(human_gse174367@meta.data$SampleID)
count_table
write.csv(count_table, file = paste0(home_dir, "/sample_counts_gse174367.csv"))
```






```{r}
# Confirming only orthologs in GEX matrix
ortholog_mat <- convert_orthologs(gene_df = row.names(human_gex_matrix),
                                    input_species = "human",
                                    output_species = "mouse",
                                    non121_strategy = "drop_both_species")

# Filtering the orthologs based on the 3250 principal component loadings
ortholog_mat <- ortholog_mat %>%
  filter(row.names(ortholog_mat) %in% row.names(principal_components))
        

# Filtering the human gex matrix based on the 3250 principal component loadings
human_gex_matrix <- human_gex_matrix %>%
  filter(row.names(human_gex_matrix) %in% ortholog_mat$input_gene)

sum(row.names(human_gex_matrix) == ortholog_mat$input_gene)

# Assigning ortholog names to human gene expression matrix
row.names(human_gex_matrix) <- row.names(ortholog_mat)

# Arrange alphabetically 
principal_components  <- principal_components %>% arrange(row.names(principal_components))

# Arrange alphabetically
human_gex_matrix <- human_gex_matrix %>%
  arrange(row.names(human_gex_matrix))

```

```{r}
# Ensure both objects are matrices/data frames with gene names as rownames
human_gex_matrix <- as.data.frame(human_gex_matrix)
principal_components <- as.data.frame(principal_components)

# Keep only shared genes, in the same order
shared_genes <- intersect(rownames(human_gex_matrix), rownames(principal_components))

human_gex_matrix <- human_gex_matrix[shared_genes, , drop = FALSE]
principal_components <- principal_components[shared_genes, , drop = FALSE]

# Optional sanity check
stopifnot(identical(rownames(human_gex_matrix), rownames(principal_components)))
```

Now the human data is put in mouse space

```{r}
sum(row.names(human_gex_matrix) == row.names(principal_components))
```



```{r}
# converting to matrix
principal_components <- as.matrix(principal_components)
human_gex_matrix <- as.matrix(human_gex_matrix)

# Ensuring equal rows, equals 3250
ncol(t(principal_components))
nrow(human_gex_matrix)

# Multiplies the transpose of the principal component matrix by the human gex matrix
multiplication_2 <- t(human_gex_matrix) %*% principal_components

# Creates new dimensional reduction on human Seurat object with the pc projections
human_gse174367[["pca_projections"]] <- CreateDimReducObject(embeddings = multiplication_2, key = "pca_proj")
```



```{r}
human_gse174367$dv <- if_else(human_gse174367$Diagnosis == "Control", 0, 1)
```

```{r}
human_gse174367 <- AddMetaData(human_gse174367, human_gse174367@reductions$pca_projections@cell.embeddings)
```


```{r}
saveRDS(human_gse174367, paste0(home_dir, "/sc_human_objects/merged_pca_on_gse174367_microglia_final_pca.RData"))
```



## Running GLMs on Projected Data
### 80% variance

```{r}
human_gse174367 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse174367_microglia_final_pca.RData"))
```

The PCs that are separative of study-specific biology include PC_1, PC_3, and PC_6 

```{r}
# Run GLM
ad.model <- glm(dv ~ pcaproj_2 +  pcaproj_4 + pcaproj_5 + 
                  pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 
                  , data = human_gse174367@meta.data, family = binomial)

```

```{r}
summary(ad.model)
```

```{r}
saveRDS(ad.model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse174367_glm_80var.RData"))
```


### AIC Optimization

```{r}
human_gse174367 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse174367_microglia_final_pca.RData"))
```

```{r}
# Run GLM
ad.model <- glm(dv ~ pcaproj_2 +  pcaproj_4 + pcaproj_5 + 
                  pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 
                  , data = human_gse174367@meta.data, family = binomial)
```

```{r message=FALSE, include=FALSE}
final_model <- stepAIC(ad.model, direction = "both")
```

```{r}
summary(final_model)
```

```{r}
saveRDS(final_model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse174367_glm_aic_optimized.RData"))
```


### PC Plots

```{r}
human_gse174367 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse174367_microglia_final_pca.RData"))
```





```{r include=FALSE}
# plot pc projections to visualize
for(i in 2:50){
p1 <- DimPlot(human_gse174367, reduction = "pca_projections", dims = c(1, i), group.by = "Diagnosis")
ggsave(filename = paste0(home_dir, "/pc_plots/pc_plot_pca_merged_on_gse174367/pc_", i, ".pdf"))
}
```


```{r}
p1 <- dittoDimPlot(human_gse174367, reduction.use = "pca_projections", dim.1 = 2, dim.2 = 5, var = "Diagnosis")
ggsave(filename = paste0(home_dir, "/figures/manuscript/pc_2_5_gse174367.pdf"), plot = p1)
```

```{r}
table(human_gse174367@meta.data$Plaque.Stage)
```





# GSE147528 (new human)
## Run PCA  for visualization

```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/gse147528_microglia_final.RData"))
```




```{r}
# Rerunning PCA again
DefaultAssay(human_gse147528) <- "RNA"
human_gse147528 <- NormalizeData(human_gse147528)
all.genes <- rownames(human_gse147528)
human_gse147528 <- ScaleData(human_gse147528, features = all.genes)
human_gex_matrix <- as.data.frame(LayerData(human_gse147528, assay = "RNA", layer = "scale.data"))
human_gse147528 <- RunPCA(human_gse147528, features = VariableFeatures(object = human_gse147528))
```



## Plotting Human GEX Matrix

```{r}
# Plotting on UMAP for visualization
human_gse147528 <- FindNeighbors(human_gse147528, dims = 1:5)
human_gse147528 <- FindClusters(human_gse147528, resolution = 0.5)
human_gse147528 <- RunUMAP(human_gse147528, dims = 1:5)
```



```{r}
p2 <- dittoDimPlot(human_gse147528, "phenotype", color.panel = c("#F8766D", "#00BFC4"))
p2
ggsave(filename= paste0(home_dir, "/figures/manuscript/umap_phenotype_human_gse147528.pdf"), width = 9, height = 7) # name the file as an EPS

```

```{r}
# Count of human microglial cells
count_table <- table(human_gse147528@meta.data$phenotype)
count_table
write.csv(count_table, file = paste0(home_dir, "/phenotype_counts_gse147528.csv"))
```



```{r}
# Count of human microglial cells
count_table <- table(human_gse147528@meta.data$orig.ident)
count_table
write.csv(count_table, file = paste0(home_dir, "/sample_counts_gse147528.csv"))
```






```{r}
# Confirming only orthologs in GEX matrix
ortholog_mat <- convert_orthologs(gene_df = row.names(human_gex_matrix),
                                    input_species = "human",
                                    output_species = "mouse",
                                    non121_strategy = "drop_both_species")

# Filtering the orthologs based on the 3250 principal component loadings
ortholog_mat <- ortholog_mat %>%
  filter(row.names(ortholog_mat) %in% row.names(principal_components))
        

# Filtering the human gex matrix based on the 3250 principal component loadings
human_gex_matrix <- human_gex_matrix %>%
  filter(row.names(human_gex_matrix) %in% ortholog_mat$input_gene)

sum(row.names(human_gex_matrix) == ortholog_mat$input_gene)

# Assigning ortholog names to human gene expression matrix
row.names(human_gex_matrix) <- row.names(ortholog_mat)

# Arrange alphabetically 
principal_components  <- principal_components %>% arrange(row.names(principal_components))

# Arrange alphabetically
human_gex_matrix <- human_gex_matrix %>%
  arrange(row.names(human_gex_matrix))

```

Now the human data is put in mouse space


```{r}
# Ensure both objects are matrices/data frames with gene names as rownames
human_gex_matrix <- as.data.frame(human_gex_matrix)
principal_components <- as.data.frame(principal_components)

# Keep only shared genes, in the same order
shared_genes <- intersect(rownames(human_gex_matrix), rownames(principal_components))

human_gex_matrix <- human_gex_matrix[shared_genes, , drop = FALSE]
principal_components <- principal_components[shared_genes, , drop = FALSE]

# Optional sanity check
stopifnot(identical(rownames(human_gex_matrix), rownames(principal_components)))
```



```{r}
# converting to matrix
principal_components <- as.matrix(principal_components)
human_gex_matrix <- as.matrix(human_gex_matrix)

# Ensuring equal rows, equals 3250
ncol(t(principal_components))
nrow(human_gex_matrix)

# Multiplies the transpose of the principal component matrix by the human gex matrix
multiplication_2 <- t(human_gex_matrix) %*% principal_components

# Creates new dimensional reduction on human Seurat object with the pc projections
human_gse147528[["pca_projections"]] <- CreateDimReducObject(embeddings = multiplication_2, key = "pca_proj")
```



```{r}
human_gse147528$dv <- if_else(human_gse147528$phenotype == "C", 0, 1)
```

```{r}
human_gse147528 <- AddMetaData(human_gse147528, human_gse147528@reductions$pca_projections@cell.embeddings)
```


```{r}
saveRDS(human_gse147528, paste0(home_dir, "/sc_human_objects/merged_pca_on_gse147528_microglia_final_pca.RData"))
```



## Running GLMs on Projected Data
### 80% variance

```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse147528_microglia_final_pca.RData"))
```

The PCs that are separative of study-specific biology include PC_1, PC_3, and PC_6 

```{r}
# Run GLM
ad.model <- glm(dv ~  pcaproj_2 +  pcaproj_4 + pcaproj_5 + 
                  pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 
                  , data = human_gse147528@meta.data, family = binomial)

```

```{r}
summary(ad.model)
```

```{r}
saveRDS(ad.model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse147528_glm_80var.RData"))
```


### AIC Optimization

```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse147528_microglia_final_pca.RData"))
```

```{r}
# Run GLM
ad.model <- glm(dv ~ pcaproj_2 +  pcaproj_4 + pcaproj_5 + 
                  pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 
                  , data = human_gse147528@meta.data, family = binomial)
```

```{r message=FALSE, include=FALSE}
final_model <- stepAIC(ad.model, direction = "both")
```

```{r}
summary(final_model)
```

```{r}
saveRDS(final_model, file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse147528_glm_aic_optimized.RData"))
```


### PC Plots

```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse147528_microglia_final_pca.RData"))
```




```{r}

human_gse147528@meta.data$BRAAK.STAGE <- if_else(human_gse147528@meta.data$orig.ident %in% c("EC1", "EC2"), "0",
                                                 if_else(human_gse147528@meta.data$orig.ident %in% c("EC4", "EC7"), "II", "VI"))
```



```{r include=FALSE}
# plot pc projections to visualize
for(i in 2:50){
  p1 <- dittoDimPlot(human_gse147528, reduction.use = "pca_projections", dim.1 = 1, dim.2 = i, var = "orig.ident")
  ggsave(filename = paste0(home_dir, "/pc_plots/pc_plot_pca_merged_on_gse147528_patient/pc_", i, ".pdf"), plot = p1)
}
```


```{r}
p1 <- dittoDimPlot(human_gse147528, reduction.use = "pca_projections", dim.1 = 2, dim.2 = 5, var = "phenotype")
ggsave(filename = paste0(home_dir, "/figures/manuscript/pc_2_5_gse147528.pdf"), plot = p1)
```





# Comparing the Performance of the Two human models


```{r}
human_gse175814 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse175814_microglia_final_pca.RData"))
```

```{r}
# Run GLM
ad.model.gse175814 <- glm(dv ~ pcaproj_1 + pcaproj_2 + pcaproj_3 + pcaproj_4 + pcaproj_5 + 
                  pcaproj_6 + pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 + pcaproj_34 + pcaproj_35 +
                  pcaproj_36
                  , data = human_gse175814@meta.data, family = binomial)
```


```{r}
human_gse147528 <- readRDS(file = paste0(home_dir, "/sc_human_objects/merged_pca_on_gse147528_microglia_final_pca.RData"))
```



```{r}
# Run GLM
ad.model.gse147528 <- glm(dv ~ pcaproj_1 + pcaproj_2 + pcaproj_3 + pcaproj_4 + pcaproj_5 + 
                  pcaproj_6 + pcaproj_7 + pcaproj_8 + pcaproj_9 + pcaproj_10 +
                  pcaproj_11 + pcaproj_12 + pcaproj_13 + pcaproj_14 + pcaproj_15 +
                  pcaproj_16 + pcaproj_17 + pcaproj_18 + pcaproj_19 + pcaproj_20 +
                  pcaproj_21 + pcaproj_22 + pcaproj_23 + pcaproj_24 + pcaproj_25 +
                  pcaproj_26 + pcaproj_27 + pcaproj_28 + pcaproj_29 + pcaproj_30 +
                  pcaproj_31 + pcaproj_32 + pcaproj_33 + pcaproj_34 + pcaproj_35 +
                  pcaproj_36
                  , data = human_gse147528@meta.data, family = binomial)

```



```{r}
summary(ad.model.gse147528)
```

```{r}
summary(ad.model.gse175814)
```








# Session Information

```{r}
sessionInfo()
```





