---
title: "AD Sparsity Analysis"
author: "Alexander Bergendorf"
date: "2024-06-05"
output: html_document
---


# Library Import

```{r}
library(Seurat)
library(mixOmics)
library(tidyverse)
library(orthogene)
library(dittoSeq)
library(elasticnet)
library(MASS)
library(pROC)
library(coop)
#home_dir <- "C:/Users/aberg/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
home_dir <- "C:/Users/Alexander/OneDrive/Desktop/Research_Brubaker/Undergraduate Research/AD TransComp-R Project Summer-Fall 2023/Brubaker Summer Research 2023"
```

# Sparsity of Dataset

```{r include=FALSE}
features <- seq(250, 15500, 250)
sparsity <- numeric(length(features))  # preallocate

mouse_gse208863 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))

merged <- merge(mouse_gse208863, mouse_gse140510, 
                add.cell.ids = c("gse208863", "gse140510"), 
                project = "merged")
combined <- JoinLayers(merged)

# Normalize once — doesn't depend on nfeatures
DefaultAssay(combined) <- "RNA"
combined <- NormalizeData(combined)

# Sparsity helper — keeps the matrix sparse, avoids as.matrix() blowup
sparsity_fn <- function(m) 1 - (Matrix::nnzero(m) / length(m))

start_time <- Sys.time()
for (idx in seq_along(features)) {
  i <- features[idx]
  
  combined <- FindVariableFeatures(combined, selection.method = "vst", nfeatures = i)
  
  data_mat <- GetAssayData(combined, assay = "RNA", layer = "data")
  data_mat <- data_mat[VariableFeatures(combined), ]
  
  sparsity[idx] <- sparsity_fn(data_mat)
  
  elapsed <- round(as.numeric(difftime(Sys.time(), start_time, units = "secs")), 1)
  message(sprintf("[%d/%d] nfeatures = %5d  |  sparsity = %.4f  |  elapsed = %ss",
                  idx, length(features), i, sparsity[idx], elapsed))
}
```



```{r}
data <- data.frame(features = features, sparsity = sparsity)

# Create scatterplot
ggplot(data, aes(x = features, y = sparsity)) +
  geom_point() +
  labs(x = "Features", y = "Sparsity") +
  ggtitle("Scatterplot of Features vs. Sparsity") +
  theme_minimal()
ggsave(filename = paste0(home_dir, "/figures/manuscript/sparsity_analysis_shared_df.pdf"), width = 7, height = 7)
write.csv(data, paste0(home_dir, "/sparsity.csv"))
```

# Performing Sparsity Analysis on PCA Data


# Regular PCA sparsity exploration

## Setup: load mouse object once, read ortholog matrix, define helper

```{r}
# Output directory for AIC/AUC results
out_dir <- "C:/Users/Alexander/OneDrive/Desktop/sparsity_analysis_results"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

# Read ortholog matrix (15102 1:1 mouse<->human orthologs across 6 datasets)
# Column "X" (unnamed in CSV) = HUMAN gene symbols (rownames from convert_orthologs)
# Column "input_gene"          = MOUSE gene symbols (input to convert_orthologs)
ortholog_df <- read.csv(paste0(home_dir, "/ortholog_matrix_IMPORTANT_6datasets.csv"),
                        stringsAsFactors = FALSE)
ortholog_df <- ortholog_df[, c("X", "input_gene")]
colnames(ortholog_df) <- c("human_gene", "mouse_gene")

# Build merged mouse object once — already-filtered microglia objects per project memory.
# Normalization is independent of nfeatures, so do it here.
mouse_gse208863 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))
merged <- merge(mouse_gse208863, mouse_gse140510,
                add.cell.ids = c("gse208863", "gse140510"),
                project = "merged")
combined_base <- JoinLayers(merged)
DefaultAssay(combined_base) <- "RNA"
combined_base <- NormalizeData(combined_base)
rm(mouse_gse208863, mouse_gse140510, merged)

# Sanity checks — run after combined_base exists
head(ortholog_df, 5)
cat("ortholog_df rows:", nrow(ortholog_df), "\n")
```





```{r}
project_and_score <- function(human, combined, ortholog_df,
                              variance_threshold = 0.80,
                              seed = 428) {

  # 1. Pull mouse PCA feature loadings (rows = mouse genes, cols = PC_1..PC_50)
  principal_components <- as.data.frame(combined@reductions$pca@feature.loadings)

  # 2. Normalize + scale human, pull GEX matrix (rows = human genes, cols = cells)
  DefaultAssay(human) <- "RNA"
  human <- NormalizeData(human)
  all.genes <- rownames(human)
  human <- ScaleData(human, features = all.genes)
  human_gex_matrix <- as.data.frame(LayerData(human, assay = "RNA", layer = "scale.data"))

  # 3. Restrict ortholog table to genes present in BOTH the mouse loadings
  #    AND the human GEX matrix. ortholog_df has columns "mouse_gene" (title-case
  #    mouse symbols) and "human_gene" (uppercase human symbols).
  ortho <- ortholog_df %>%
    filter(mouse_gene %in% rownames(principal_components),
           human_gene %in% rownames(human_gex_matrix))

  # 4. Subset and align both matrices to that shared, ordered ortholog set
  principal_components <- principal_components[ortho$mouse_gene, , drop = FALSE]
  human_gex_matrix     <- human_gex_matrix[ortho$human_gene, , drop = FALSE]

  # Rename human GEX rows to mouse symbols so projection happens in mouse gene space
  rownames(human_gex_matrix) <- ortho$mouse_gene

  # Sanity check: rows must match exactly, in order
  stopifnot(identical(rownames(principal_components), rownames(human_gex_matrix)))

  # 5. Project: t(human_GEX) %*% loadings  ->  cells x PCs
  #    (matches the formulation used to build merged_pca_on_gse175814_microglia_final_pca.RData)
  principal_components <- as.matrix(principal_components)
  human_gex_matrix     <- as.matrix(human_gex_matrix)
  embeddings_mat <- t(human_gex_matrix) %*% principal_components   # cells x PCs

  # 6. Attach projection as a DimReduc on the human object
  human[["pca_projections"]] <- CreateDimReducObject(
    embeddings = embeddings_mat,
    key = "pcaproj_",
    assay = DefaultAssay(human)
  )

  # Add per-PC embeddings as metadata columns named pcaproj_1, pcaproj_2, ...
  proj_embeddings <- human@reductions$pca_projections@cell.embeddings
  pc_colnames <- colnames(proj_embeddings)
  human <- AddMetaData(human, proj_embeddings, col.name = pc_colnames)

  # 7. Build response variable from `phenotype` (AD = 1, C = 0)
  human@meta.data$dv <- as.integer(human@meta.data$phenotype == "AD")

  # 8. Variance explained by mouse PCs — keep PCs whose cum_var is strictly below
  #    the threshold (matches original sparsity-analysis methodology).
  stdev <- combined@reductions$pca@stdev
  variance <- stdev^2 / sum(stdev^2)
  cum_var <- cumsum(variance)
  PCs <- max(1, sum(cum_var < variance_threshold))

  # 9. Train/test split, GLM with stepAIC, AUC on held-out
  set.seed(seed)
  pc_keep   <- pc_colnames[seq_len(PCs)]
  model_df  <- human@meta.data[, c(pc_keep, "dv"), drop = FALSE]
  model_df  <- model_df[complete.cases(model_df), , drop = FALSE]

  train_idx <- sample(c(TRUE, FALSE), nrow(model_df),
                      replace = TRUE, prob = c(0.8, 0.2))
  train_df  <- model_df[train_idx, , drop = FALSE]
  test_df   <- model_df[!train_idx, , drop = FALSE]

  ad.model     <- glm(dv ~ ., data = train_df, family = binomial)
  final_model  <- MASS::stepAIC(ad.model, direction = "both", trace = FALSE)

  predictions  <- predict(final_model, newdata = test_df, type = "response")
  roc_obj      <- pROC::roc(test_df$dv, predictions, quiet = TRUE)

  list(auc = as.numeric(roc_obj$auc),
       aic = final_model$aic,
       n_pcs = PCs)
}
```



## Run for GSE175814

```{r}
features <- seq(250, 15250, 250)
auc_vec_175814 <- list()
aic_vec_175814 <- list()

human_gse175814 <- readRDS(paste0(home_dir, "/sc_human_objects/gse175814_microglia_final.RData"))

for (i in features) {
  message("[GSE175814] features = ", i)

  # Re-derive PCA on the merged mouse object at this nfeatures setting
  combined <- combined_base
  combined <- FindVariableFeatures(combined, selection.method = "vst", nfeatures = i, verbose = FALSE)
  all.genes <- rownames(combined)
  combined <- ScaleData(combined, features = all.genes, verbose = FALSE)
  combined <- RunPCA(combined,
                     features = VariableFeatures(object = combined),
                     verbose = FALSE)

  res <- project_and_score(human = human_gse175814,
                           combined = combined,
                           ortholog_df = ortholog_df)

  auc_vec_175814[[as.character(i)]] <- res$auc
  aic_vec_175814[[as.character(i)]] <- res$aic

  saveRDS(auc_vec_175814, file.path(out_dir, "auc_pca_gse175814.RData"))
  saveRDS(aic_vec_175814, file.path(out_dir, "aic_pca_gse175814.RData"))
}

rm(human_gse175814)
```




## Run for GSE147528

```{r}
features <- seq(250, 15250, 250)
auc_vec_147528 <- list()
aic_vec_147528 <- list()

human_gse147528 <- readRDS(paste0(home_dir, "/sc_human_objects/gse147528_microglia_final.RData"))

for (i in features) {
  message("[GSE147528] features = ", i)

  combined <- combined_base
  combined <- FindVariableFeatures(combined, selection.method = "vst", nfeatures = i)
  all.genes <- rownames(combined)
  combined <- ScaleData(combined, features = all.genes)
  combined <- RunPCA(combined,
                     features = VariableFeatures(object = combined),
                     verbose = FALSE)

  res <- project_and_score(human = human_gse147528,
                           combined = combined,
                           ortholog_df = ortholog_df)

  auc_vec_147528[[as.character(i)]] <- res$auc
  aic_vec_147528[[as.character(i)]] <- res$aic

  saveRDS(auc_vec_147528, file.path(out_dir, "auc_pca_gse147528.RData"))
  saveRDS(aic_vec_147528, file.path(out_dir, "aic_pca_gse147528.RData"))
}

rm(human_gse147528)
```



## Combined plots: GSE175814 vs GSE147528

```{r}
# Build a combined tidy data frame
plot_df_combined <- rbind(
  data.frame(
    features = as.numeric(names(auc_vec_175814)),
    AUC      = unlist(auc_vec_175814, use.names = FALSE),
    AIC      = unlist(aic_vec_175814, use.names = FALSE),
    Dataset  = "GSE175814"
  ),
  data.frame(
    features = as.numeric(names(auc_vec_147528)),
    AUC      = unlist(auc_vec_147528, use.names = FALSE),
    AIC      = unlist(aic_vec_147528, use.names = FALSE),
    Dataset  = "GSE147528"
  )
)
plot_df_combined <- plot_df_combined[order(plot_df_combined$Dataset,
                                           plot_df_combined$features), ]

# AIC: both datasets on one plot
p_aic_combined <- ggplot(plot_df_combined,
                         aes(x = features, y = AIC, color = Dataset)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(x = "nfeatures", y = "AIC",
       title = "PCA sparsity exploration — AIC",
       color = "Human dataset") +
  theme_classic()

# AUC: both datasets on one plot
p_auc_combined <- ggplot(plot_df_combined,
                         aes(x = features, y = AUC, color = Dataset)) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(x = "nfeatures", y = "AUC",
       title = "PCA sparsity exploration — AUC",
       color = "Human dataset") +
  theme_classic()

# Save PDFs alongside the per-dataset outputs
ggsave(file.path(out_dir, "aic_pca_combined.pdf"),
       plot = p_aic_combined, width = 7.5, height = 5)
ggsave(file.path(out_dir, "auc_pca_combined.pdf"),
       plot = p_auc_combined, width = 7.5, height = 5)

# Print to notebook
p_aic_combined
p_auc_combined
```

## Sparse PCA sparsity exploration

```{r}
# Output directory for AIC/AUC results
out_dir <- "C:/Users/Alexander/OneDrive/Desktop/sparsity_analysis_results"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

# Read ortholog matrix (15102 1:1 mouse<->human orthologs across 6 datasets)
# Column "X" (unnamed in CSV) = HUMAN gene symbols (rownames from convert_orthologs)
# Column "input_gene"          = MOUSE gene symbols (input to convert_orthologs)
ortholog_df <- read.csv(paste0(home_dir, "/ortholog_matrix_IMPORTANT_6datasets.csv"),
                        stringsAsFactors = FALSE)
ortholog_df <- ortholog_df[, c("X", "input_gene")]
colnames(ortholog_df) <- c("human_gene", "mouse_gene")

# Build merged mouse object once — already-filtered microglia objects per project memory.
# Normalization is independent of nfeatures, so do it here.
mouse_gse208863 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse208683_microglia_final.RData"))
mouse_gse140510 <- readRDS(paste0(home_dir, "/sc_mouse_objects/gse140510_microglia_final.RData"))
merged <- merge(mouse_gse208863, mouse_gse140510,
                add.cell.ids = c("gse208863", "gse140510"),
                project = "merged")
combined_base <- JoinLayers(merged)
DefaultAssay(combined_base) <- "RNA"
combined_base <- NormalizeData(combined_base)
rm(mouse_gse208863, mouse_gse140510, merged)

# Sanity checks — run after combined_base exists
head(ortholog_df, 5)
cat("ortholog_df rows:", nrow(ortholog_df), "\n")
```

## Helper Functions

```{r}
prepare_human <- function(human, ortholog_df) {
  DefaultAssay(human) <- "RNA"
  human <- NormalizeData(human, verbose = FALSE)
  human <- ScaleData(human, features = rownames(human), verbose = FALSE)
  human_gex_matrix <- LayerData(human, assay = "RNA", layer = "scale.data")

  ortho_human_subset <- ortholog_df[ortholog_df$human_gene %in% rownames(human_gex_matrix), ]

  dv <- as.integer(human@meta.data$phenotype == "AD")
  names(dv) <- rownames(human@meta.data)

  list(
    gex_matrix = human_gex_matrix,
    ortho      = ortho_human_subset,
    dv         = dv,
    cell_ids   = colnames(human_gex_matrix)
  )
}

make_split <- function(n_cells, seed = 428) {
  set.seed(seed)
  sample(c(TRUE, FALSE), n_cells, replace = TRUE, prob = c(0.8, 0.2))
}

fit_mouse_spca <- function(combined_base, nfeatures, keepX, ncomp = 35) {

  effective_keepX <- min(keepX, nfeatures)

  combined <- combined_base
  combined <- FindVariableFeatures(combined, selection.method = "vst",
                                   nfeatures = nfeatures, verbose = FALSE)
  vf <- VariableFeatures(combined)

  # Scale ONLY variable features — saves time vs scaling all genes
  combined <- ScaleData(combined, features = vf, verbose = FALSE)

  scaled <- LayerData(combined, assay = "RNA", layer = "scale.data")[vf, , drop = FALSE]

  bad_rows <- !is.finite(rowSums(scaled))
  if (any(bad_rows)) {
    message("    dropping ", sum(bad_rows), " features with non-finite values")
    scaled <- scaled[!bad_rows, , drop = FALSE]
  }

  mouse_mat <- t(scaled)
  if (!is.matrix(mouse_mat)) mouse_mat <- as.matrix(mouse_mat)
  effective_keepX <- min(effective_keepX, ncol(mouse_mat))

  mouse_embeddings <- tryCatch(
    mixOmics::spca(mouse_mat, ncomp = ncomp,
                   center = FALSE, scale = FALSE,
                   keepX = rep(effective_keepX, ncomp)),
    error = function(e) {
      message("    spca failed at ncomp=", ncomp, ": ", conditionMessage(e))
      message("    retrying with ncomp = 25")
      mixOmics::spca(mouse_mat, ncomp = 25,
                     center = FALSE, scale = FALSE,
                     keepX = rep(effective_keepX, 25))
    }
  )

  combined[["spca"]] <- CreateDimReducObject(
    embeddings = mouse_embeddings$x,
    loadings   = mouse_embeddings$rotation,
    key        = "spca_",
    assay      = DefaultAssay(combined)
  )

  combined
}

project_and_score_spca <- function(human_pre, combined, train_idx) {

  loadings <- combined@reductions$spca@feature.loadings

  ortho <- human_pre$ortho[human_pre$ortho$mouse_gene %in% rownames(loadings), ]

  loadings_sub <- loadings[ortho$mouse_gene, , drop = FALSE]
  gex_sub      <- human_pre$gex_matrix[ortho$human_gene, , drop = FALSE]

  stopifnot(nrow(loadings_sub) == nrow(gex_sub))

  embeddings_mat <- crossprod(gex_sub, loadings_sub)

  pc_names <- paste0("spcaproj_", seq_len(ncol(embeddings_mat)))
  colnames(embeddings_mat) <- pc_names

  model_df <- data.frame(as.matrix(embeddings_mat),
                         dv = human_pre$dv[rownames(embeddings_mat)],
                         check.names = FALSE)
  model_df <- model_df[complete.cases(model_df), , drop = FALSE]

  train_df <- model_df[train_idx, , drop = FALSE]
  test_df  <- model_df[!train_idx, , drop = FALSE]

  ad.model    <- glm(dv ~ ., data = train_df, family = binomial)
  final_model <- MASS::stepAIC(ad.model, direction = "both", trace = FALSE)

  predictions <- predict(final_model, newdata = test_df, type = "response")
  roc_obj     <- pROC::roc(test_df$dv, predictions, quiet = TRUE)

  list(auc = as.numeric(roc_obj$auc),
       aic = final_model$aic,
       n_pcs = ncol(embeddings_mat))
}
```

## Run for GSE175814

```{r}
features   <- seq(750, 15250, 250)
keepX_vals <- c(100, 200, 500)

human_gse175814  <- readRDS(paste0(home_dir, "/sc_human_objects/gse175814_microglia_final.RData"))
human_pre_175814 <- prepare_human(human_gse175814, ortholog_df)
train_idx_175814 <- make_split(ncol(human_pre_175814$gex_matrix))
rm(human_gse175814)

for (kx in keepX_vals) {
  message("[GSE175814] keepX = ", kx)
  auc_vec <- list()
  aic_vec <- list()

  for (i in features) {
    message("  features = ", i)

    combined <- fit_mouse_spca(combined_base, nfeatures = i, keepX = kx, ncomp = 35)
    res <- project_and_score_spca(human_pre_175814, combined, train_idx_175814)

    auc_vec[[as.character(i)]] <- res$auc
    aic_vec[[as.character(i)]] <- res$aic

    saveRDS(auc_vec, file.path(out_dir,
            paste0("auc_spca_keepX", kx, "_gse175814.RData")))
    saveRDS(aic_vec, file.path(out_dir,
            paste0("aic_spca_keepX", kx, "_gse175814.RData")))
  }
}

rm(human_pre_175814)
```

## Run for GSE147528


## Combined sPCA plots: GSE175814 vs GSE147528 across keepX values

```{r}
keepX_vals <- c(100, 200, 500)

# Helper: load a saved list and convert to long-format rows
load_run <- function(metric, kx, dataset) {
  path <- file.path(out_dir,
                    paste0(metric, "_spca_keepX", kx, "_", dataset, ".RData"))
  if (!file.exists(path)) return(NULL)
  vals <- readRDS(path)
  data.frame(
    features = as.numeric(names(vals)),
    value    = unlist(vals, use.names = FALSE),
    keepX    = kx,
    Dataset  = toupper(dataset),
    metric   = metric
  )
}

# Load every (metric, keepX, dataset) combination and stack
plot_df_spca <- do.call(rbind, c(
  lapply(keepX_vals, function(kx) load_run("auc", kx, "gse175814")),
  lapply(keepX_vals, function(kx) load_run("auc", kx, "gse147528")),
  lapply(keepX_vals, function(kx) load_run("aic", kx, "gse175814")),
  lapply(keepX_vals, function(kx) load_run("aic", kx, "gse147528"))
))

# keepX as a factor with ordered levels for clean legend ordering
plot_df_spca$keepX <- factor(plot_df_spca$keepX, levels = keepX_vals)

# Split for separate AIC and AUC plots
df_aic <- subset(plot_df_spca, metric == "aic")
df_auc <- subset(plot_df_spca, metric == "auc")

# AIC: faceted by Dataset, lines colored by keepX
p_aic_spca <- ggplot(df_aic,
                     aes(x = features, y = value, color = keepX)) +
  geom_line() +
  geom_point(size = 1.2) +
  facet_wrap(~ Dataset, scales = "free_y") +
  labs(x = "nfeatures", y = "AIC",
       title = "sPCA sparsity exploration — AIC",
       color = "keepX") +
  theme_classic()

# AUC: faceted by Dataset, lines colored by keepX
p_auc_spca <- ggplot(df_auc,
                     aes(x = features, y = value, color = keepX)) +
  geom_line() +
  geom_point(size = 1.2) +
  facet_wrap(~ Dataset) +
  labs(x = "nfeatures", y = "AUC",
       title = "sPCA sparsity exploration — AUC",
       color = "keepX") +
  theme_classic()

# Save PDFs
ggsave(file.path(out_dir, "aic_spca_combined.pdf"),
       plot = p_aic_spca, width = 11, height = 5)
ggsave(file.path(out_dir, "auc_spca_combined.pdf"),
       plot = p_auc_spca, width = 11, height = 5)

# Print to notebook
p_aic_spca
p_auc_spca
```

# Session Information

```{r}
sessionInfo()
```

